---
title: "IL6 Serum Biologics"
output: html_notebook
---

## Libraries

```{r}
library(cowplot)
library(ggplot2)
library(limma)
library(dplyr)
library(plotly)
library(ggrepel)
library(svglite)
library(ComplexHeatmap)
library(tidyverse)
```

## Import dataframes

```{r}
all_peptides <- read.csv("Input/250121_T1D_Serum_EV_Treatment.csv",
                        header=TRUE, 
                        na = "#N/A") %>%
  # This protein has significant peptide overlap with Siltuximab
  # It is removed here to avoid spurious results/incorrect quantitation
  # Open to other ideas
  filter(!grepl("HV145",Protein)) 



meta <- read.csv("Input/250125_meta.csv",
                 stringsAsFactors = FALSE)
```

### Roll up into proteins

```{r}
protein_data <- all_peptides %>%
  group_by(Protein) %>%
  summarise(across(7:ncol(all_peptides)-1, sum, na.rm = TRUE))
protein_data <- protein_data %>%
  mutate(across(-Protein, ~ replace(., . == 0, NA)))
```
### QC - protein and peptide counts
```{r}
counts <- read.csv("Input/peptide_protein_count.csv") %>%
  select(!contains("DPRef"))


library(dplyr)
library(tidyr)
library(ggplot2)
library(cowplot)
library(stringr)

if (!dir.exists("Output/barplots")) dir.create("Output/barplots", recursive = TRUE)

peptide_cols <- grep("Count\\.Distinct\\.Peptide$", colnames(counts), value = TRUE)
protein_cols <- grep("Count\\.Distinct\\.Protein$", colnames(counts), value = TRUE)

## --- Peptide count barplot ---------------------------------------------

df_peptide <- counts %>%
  select(all_of(peptide_cols)) %>%
  pivot_longer(cols = everything(), names_to = "sample", values_to = "peptide_count") %>%
  mutate(sample_label = sub("_.*", "", sample))

df_peptide$sample_label <- factor(df_peptide$sample_label, levels = unique(df_peptide$sample_label))

p_peptide <- ggplot(df_peptide, aes(x = sample_label, y = peptide_count)) +
  geom_col(fill = "grey70", color = "black", width = 0.7) +
  theme_cowplot() +
  labs(title = "Distinct peptide count by sample",
       y = "Distinct peptide count",
       x = NULL) +
  theme(
    panel.background = element_rect(fill = "white", color = NA),
    plot.background = element_rect(fill = "white", color = NA),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8)
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))

ggsave("Output/barplots/peptide_count_by_sample.pdf", plot = p_peptide, width = 12, height = 5)

## --- Protein count barplot ----------------------------------------------

df_protein <- counts %>%
  select(all_of(protein_cols)) %>%
  pivot_longer(cols = everything(), names_to = "sample", values_to = "protein_count") %>%
  mutate(sample_label = sub("_.*", "", sample))

df_protein$sample_label <- factor(df_protein$sample_label, levels = unique(df_protein$sample_label))

p_protein <- ggplot(df_protein, aes(x = sample_label, y = protein_count)) +
  geom_col(fill = "grey70", color = "black", width = 0.7) +
  theme_cowplot() +
  labs(title = "Distinct protein count by sample",
       y = "Distinct protein count",
       x = NULL) +
  theme(
    panel.background = element_rect(fill = "white", color = NA),
    plot.background = element_rect(fill = "white", color = NA),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8)
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))

ggsave("Output/barplots/protein_count_by_sample.pdf", plot = p_protein, width = 12, height = 5)
```

```{r}
df_counts_agg <- bind_rows(
  counts %>%
    select(all_of(peptide_cols)) %>%
    pivot_longer(cols = everything(), names_to = "sample", values_to = "count") %>%
    mutate(metric = "Peptide"),
  counts %>%
    select(all_of(protein_cols)) %>%
    pivot_longer(cols = everything(), names_to = "sample", values_to = "count") %>%
    mutate(metric = "Protein")
)

df_counts_agg$metric <- factor(df_counts_agg$metric, levels = c("Peptide", "Protein"))

p_counts_agg <- ggplot(df_counts_agg, aes(x = metric, y = count, fill = metric)) +
  stat_summary(fun = mean,
               geom = "bar",
               color = "black",
               width = 0.7,
               alpha = 0.5) +
  stat_summary(fun.data = mean_sdl,
               fun.args = list(mult = 1),
               geom = "errorbar", width = 0.2) +
  geom_quasirandom(width = 0.2, size = 2, shape = 21, color = "black") +
  stat_summary(fun = mean,
               geom = "text",
               aes(label = paste("mean\n",round(after_stat(y)))),
               vjust = -0.3, size = 3.5, fontface = "bold") +
  scale_fill_manual(values = c("Peptide" = "#F2C438", "Protein" = "#9c5ae2")) +
  theme_cowplot() +
  labs(title = NULL,
       y = "Count",
       x = NULL) +
  theme(
    panel.background = element_rect(fill = "white", color = NA),
    plot.background = element_rect(fill = "white", color = NA),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    legend.position = "none"
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.25)))

ggsave("Output/barplots/peptide_vs_protein_count_aggregate.pdf", plot = p_counts_agg, width = 2.5 , height = 2.5)
```


### Data manipulation, mAb peptide-level

#### Siltuximab

-   Results in "s.peptide_data", a dataframe with only siltuximab peptides in siltuximab treatment
-   Writes siltux_peptides.csv for plotting

```{r}
df <- all_peptides %>%
  filter(grepl("siltux", Protein, ignore.case=TRUE))

df$Protein_peptide <- paste0(df$Protein,"_",df$Peptide.Modified.Sequence)

row.names(df) <- df$Protein_peptide

Siltux_baseline_sequence <- sprintf("S%02d",seq(1,19,by=2))
Siltux_treat_sequence <- sprintf("S%02d",seq(2,20,by=2))

Tocil_baseline_sequence <- sprintf("T%02d",seq(1,19,by=2))
Tocil_treat_sequence <- sprintf("T%02d",seq(2,20,by=2))

Siltux_baseline_peptides <- select(df, starts_with(Siltux_baseline_sequence))
Siltux_treat_peptides <- select(df, starts_with(Siltux_treat_sequence))

s.peptide_data <- cbind(Siltux_baseline_peptides, Siltux_treat_peptides)

row.names(s.peptide_data) <- row.names(df)

write.csv(s.peptide_data, "Output/siltux_peptides.csv")
```

#### Tocilizumab

-   Results in "t.peptide_data", a dataframe with only siltuximab peptides in siltuximab treatment
-   Writes tocilizumab_peptides.csv for plotting

```{r}
df <- all_peptides %>%
  filter(grepl("tocil", Protein, ignore.case=TRUE))

df$Protein_peptide <- paste0(df$Protein,"_",df$Peptide.Modified.Sequence)

row.names(df) <- df$Protein_peptide

#Siltux_baseline_sequence <- sprintf("S%02d",seq(1,19,by=2))
#Siltux_treat_sequence <- sprintf("S%02d",seq(2,20,by=2))

#Tocil_baseline_sequence <- sprintf("T%02d",seq(1,19,by=2))
#Tocil_treat_sequence <- sprintf("T%02d",seq(2,20,by=2))

Tocil_baseline_peptides <- select(df, starts_with(Tocil_baseline_sequence))
Tocil_treat_peptides <- select(df, starts_with(Tocil_treat_sequence))

t.peptide_data <- cbind(Tocil_baseline_peptides, Tocil_treat_peptides)

row.names(t.peptide_data) <- row.names(df)

write.csv(t.peptide_data,"Output/tocilizumab_peptides.csv")
```

### Rolling up mAbs into protein-level from a select few peptides

-   Requires an input, manually-filtered set of mAb peptides
    -   These were peptides belonging to the mAbs which satisfied the two following conditions:
        -   Were unique to the mAb (not found in other, endogenous Igs)
            -   NOTE: one peptide quantifying siltuximab (SEDTA...) overlaps with HV145 but is seemingly mostly coming from siltuximab given that it correlates with other siltuximab peptides. Consider removing this peptide from Ab quantitation
        -   Significantly increased in treatment (t-tests performed in Prism)
    -   Does not require running the above two sections

```{r}
quant_peptides <- read.csv("Input/250124_treatment_peptides.csv", header = FALSE)

all_peptides_siltux <- all_peptides %>%
  filter(grepl("siltux", Protein, ignore.case=TRUE))

all_peptides_siltux$Protein_peptide <- 
  paste0(all_peptides_siltux$Protein,"_", 
  all_peptides_siltux$Peptide.Modified.Sequence)

query <- quant_peptides
rownames(all_peptides_siltux) <- all_peptides_siltux$Protein_peptide

all_rownames <- rownames(all_peptides_siltux)

query <- query[[1]]
matches <- sapply(all_rownames, function(protein) {
any(sapply(query, function(name) grepl(name, protein)))
})
filtered_query <- all_peptides_siltux[matches, ]

siltux_good_peptides <- filtered_query

siltux_quant <- siltux_good_peptides %>%
  summarise(across(where(is.numeric), sum, na.rm=TRUE))

siltux_quant$Protein = "Siltux_filtered"

all_peptides_tocil <- all_peptides %>%
  filter(grepl("tocil", Protein, ignore.case=TRUE))

all_peptides_tocil$Protein_peptide <- paste0(all_peptides_tocil$Protein,"_", all_peptides_tocil$Peptide.Modified.Sequence)



query <- quant_peptides
rownames(all_peptides_tocil) <- all_peptides_tocil$Protein_peptide

all_rownames <- rownames(all_peptides_tocil)

query <- query[[1]]
matches <- sapply(all_rownames, function(protein) {
any(sapply(query, function(name) grepl(name, protein)))
})
filtered_query <- all_peptides_tocil[matches, ]

tocil_good_peptides <- filtered_query

tocil_quant <- tocil_good_peptides %>%
  summarise(across(where(is.numeric), sum, na.rm=TRUE))

tocil_quant$Protein = "Tocil_filtered"

protein_data <- rbind(protein_data, tocil_quant) %>%
  rbind(., siltux_quant)
```

### Data manipulation, protein-level, with quantitation of tocilizumab and situximab included

```{r}
all_proteins <- protein_data

row.names(all_proteins) <- all_proteins$Protein

Siltux_baseline_sequence <- sprintf("S%02d",seq(1,19,by=2))
Siltux_treat_sequence <- sprintf("S%02d",seq(2,20,by=2))

Tocil_baseline_sequence <- sprintf("T%02d",seq(1,19,by=2))
Tocil_treat_sequence <- sprintf("T%02d",seq(2,20,by=2))

Siltux_baseline <- select(all_proteins, starts_with(Siltux_baseline_sequence))
Siltux_treat <- select(all_proteins, starts_with(Siltux_treat_sequence))

s.protein_data <- cbind(Siltux_baseline, Siltux_treat)

row.names(s.protein_data) <- row.names(all_proteins)

#write.csv(s.protein_data, "250125_siltux.csv")

Tocil_baseline <- select(all_proteins, starts_with(Tocil_baseline_sequence))
Tocil_treat <- select(all_proteins, starts_with(Tocil_treat_sequence))

t.protein_data <- cbind(Tocil_baseline, Tocil_treat)

row.names(t.protein_data) <- row.names(all_proteins)

#write.csv(t.protein_data,"250121_tocil.csv")

all_samples <- cbind(s.protein_data, t.protein_data)

```

### Data manipulation, peptide-level

```{r}
all_peptides_unique <- all_peptides %>%
  distinct(Peptide, .keep_all = TRUE)
  

row.names(all_peptides_unique) <- all_peptides_unique$Peptide

Siltux_baseline_sequence <- sprintf("S%02d",seq(1,19,by=2))
Siltux_treat_sequence <- sprintf("S%02d",seq(2,20,by=2))

Tocil_baseline_sequence <- sprintf("T%02d",seq(1,19,by=2))
Tocil_treat_sequence <- sprintf("T%02d",seq(2,20,by=2))

Siltux_baseline <- select(all_peptides_unique, starts_with(Siltux_baseline_sequence))
Siltux_treat <- select(all_peptides_unique, starts_with(Siltux_treat_sequence))

s.peptide_data <- cbind(Siltux_baseline, Siltux_treat)

row.names(s.peptide_data) <- row.names(all_peptides_unique)

#write.csv(s.protein_data, "250125_siltux.csv")

Tocil_baseline <- select(all_peptides_unique, starts_with(Tocil_baseline_sequence))
Tocil_treat <- select(all_peptides_unique, starts_with(Tocil_treat_sequence))

t.peptide_data <- cbind(Tocil_baseline, Tocil_treat)

row.names(t.peptide_data) <- row.names(all_peptides_unique)

#write.csv(t.protein_data,"250121_tocil.csv")

all_samples_peptides<- cbind(s.peptide_data, t.peptide_data)
#write.csv(all_samples_peptides, "Output/2511.csv")
```

## Protein-level equalize medians

```{r}
df <- all_samples

i=1


info <- data.frame()
for(i in 1:ncol(df)){
  name <- colnames(df)[i]
  intensities <- df[,i]
  median <- median(intensities, na.rm=TRUE)
  var <- var(intensities, na.rm=TRUE)
  df_add <- data.frame(name, median, var)
  info <- rbind(info, df_add)
}

med_med <- median(info$median)

ggplot(info, aes(x = name, y = median)) +
  geom_col(fill = "steelblue") +
  theme_cowplot() +
  labs(x = "Name", y = "Median") +
  theme(axis.text.x = element_text(angle=90, hjust=0.5, vjust=0.5, size=10))

normed <- data.frame(row.names = row.names(df))
i=1

for(i in 1:ncol(df)){
  normed[,i]=df[,i]/info$median[i]*med_med
}

# MM feedback - multiply by a median of medians to get magnitude of signal back to where it was -- current problem is some signals look artifically tiny if they are 0.00001 of the median

colnames(normed) <- colnames(df)

normed_info <- data.frame()
for(i in 1:ncol(normed)){
  name <- colnames(normed)[i]
  intensities <- normed[,i]
  median <- median(intensities, na.rm=TRUE)
  var <- var(intensities, na.rm=TRUE)
  df_add <- data.frame(name, median, var)
  normed_info <- rbind(normed_info, df_add)
}



ggplot(normed_info, aes(x = name, y = median)) +
  geom_col(fill = "steelblue") +
  theme_cowplot() +
  labs(x = "Name", y = "Median") +
  theme(axis.text.x = element_text(angle=90, hjust=0.5, vjust=0.5, size=10))
```

#### Export protein-level equalize median normalized dataframe

```{r}
normed_gn <- normed %>%
  mutate(Reference = row.names(.)) %>%
  left_join(
    all_peptides %>% 
      select(Protein, Protein.Accession) %>% 
      distinct(Protein, .keep_all = TRUE),
    by = c("Reference" = "Protein")
  )%>%
  mutate(
    Protein.Accession = if_else(is.na(Protein.Accession) | Protein.Accession == "", Reference, Protein.Accession
  ))

#From uniprot website, convert accession number to proper gene name
uniprot_gene_name <- read.csv("Input/Uniprot_to_gene_name.csv")

normed_gn <- normed_gn %>%
  mutate(
    Accession_primary = trimws(sub("/.*$", "", Protein.Accession))
  ) %>%
  left_join(
    uniprot_gene_name %>% select(Entry, GeneName),
    by = c("Accession_primary" = "Entry")
  ) %>%
  mutate(
    GeneName = if_else(is.na(GeneName) | GeneName == "", Protein.Accession, GeneName)
  ) %>%
  select(Reference, Protein.Accession, GeneName, everything())

row.names(normed_gn) <- normed_gn$GeneName

write.csv(normed_gn,"Output/251125_normed_proteins.csv")
```

### Barplots of interest

```{r}

library(ggtext)
library(ggbeeswarm)
library(ggsci)
library(cowplot)
normed_gn <- read.csv("Output/251125_normed_proteins.csv")

var <- c(rep(c("siltuximab baseline", "siltuximab treatment", "tocilizumab baseline","tocilizumab treatment"), each=10))

#txt <- "PSMD6,PSMD13,PSMC1,PSMD2,PSMD8,IKBKB,PSMC5,PSMD12,PSMC2,PSMD14,PSMC3,PSMB5,PSMC4,UBB,NFKB1,PSMD1,PSMC6,PSMB6,MAP3K7,PSMA5,PSMD11,PSMB4,PSMB3,PSMA4,PSMB7,PSMA7,PSMA2,PSMB2,PSMA1,PSMA3,PSMA6"
txt <- "PPBP"

selected <- strsplit(txt, ",")[[1]]

selected[[1]]

for(i in 1:length(selected)){

conditions <- var


df_plot <- normed_gn %>%
  filter(GeneName == selected[[i]]) %>%
  select(-Reference, -Protein.Accession, -Accession_primary)

if (nrow(df_plot) == 0) {
  message("Skipping missing gene: ", selected[[i]])
  next
}

df_long <- df_plot %>%
  pivot_longer(col = 2:41,
               names_to = "condition",
               values_to = "normalized abundance") %>%
  mutate(condition=conditions)

df_long$condition <- factor(df_long$condition, levels = c("siltuximab baseline", "siltuximab treatment", "tocilizumab baseline","tocilizumab treatment"))

#description_wrapped <- stringr::str_wrap(df_long$Description[[1]], width = 10)
#description_wrapped <- gsub("\n", "<br>", description_wrapped)


ggplot(df_long, aes(x = condition, y = `normalized abundance`, fill = condition)) +
  stat_summary(fun = mean, 
               geom = "bar", 
               color = "black", 
               width = 0.7, 
               alpha=0.5) +        
  stat_summary(fun.data = mean_sdl,
               fun.args = list(mult=1), 
               geom = "errorbar", width = 0.2) +            
  geom_quasirandom(width = 0.2 , size = 2, shape = 21, color = "black") +
  scale_fill_viridis_d() +
  #scale_fill_brewer(palette = "Set1") +
  theme_cowplot() +
  labs(
    title  = df_long$GeneName[[1]],
    #subtitle = stringr::str_wrap(df_long$Description[[1]], width = 35),
    y = "normalized abundance",
    x = NULL
  ) +
  theme(
    panel.background = element_rect(fill = "white", color = NA),
  plot.background = element_rect(fill = "white", color = NA),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1))+
  scale_y_continuous(expand=expansion(mult=c(0,0.1)))

ggsave(paste0("Output/barplots/",df_long$GeneName[[1]],".png"), width=4, height=5)
}
```

## Limma, volcano plots

### Siltuximab Limma

```{r}

#Select only siltuximab treatment samples
num.df <- normed_gn[1:20]
#rownames(num.df) <-s.protein_data[,1]

#Removing 0's because they don't log2
num.df[num.df==0] <- NA

#make the data a matrix of values
mat <- data.matrix(num.df)
#Limma expects a log2 transform
l2m <- log2(mat)

#Extracting conditions from metadata file
patient <- factor(meta$Patient)
treat <- factor(meta$Condition, level=c('SB','ST'))

design <- model.matrix(~patient+treat)

fit <- lmFit(l2m, design)
fits <- eBayes(fit)

#contrast.matrix <- makeContrasts(
  #Siltux_Treatment_vs_Baseline = ST - SB,
 # levels = design
#)

s.results <- topTable(fits, coef = "treatST", adjust="BH", n=Inf)
c.s.results <- s.results
c.s.results["-log10_adj_p-val"] <- -log10(c.s.results[5])

c.s.results <- c.s.results %>% mutate(gene = sub(".*\\|(.*?)_.*", "\\1", row.names(c.s.results)))
sig_s_results <- c.s.results[which(c.s.results$adj.P.Val<0.05),]

#write.csv(c.s.results, "250219 Siltux treatment vs baseline PAIRED extra_median normalized imputed.csv", row.names = TRUE)


```

```{r}
library(ggbreak)

ggplot(c.s.results, aes(x = logFC, y = `-log10_adj_p-val`)) + 
  geom_point(color = "grey80") + 
  theme_cowplot() + 
  geom_point(data = sig_s_results, aes(x = logFC, y = `-log10_adj_p-val`), color = "black") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black") +
  geom_text_repel(
    data = sig_s_results,
    aes(x = logFC, y = `-log10_adj_p-val`, label = gene),
    color = "black",
    size = 3,
    box.padding = 0.3,
    segment.color = "black",
    max.overlaps = 25,
    fill = "white",     # gives white background bubble
    min.segment.length = 0,
    bg.color = "white", # for ggrepel ≥0.9.3, use bg.color/bg.r
    bg.r = 0.15         # rounded corners
  ) +

   
  scale_x_continuous(limits = c(-4, 4), breaks = seq(-4, 4, 1)) + 
  scale_y_continuous(limits = c(0, 4.5), breaks = seq(0, 4.5, 0.5)) + 
  scale_y_break(c(2.2,3.4)) +

  labs(x="Siltuximab log2(treatment/baseline)", y="-log10(adj. P-value)")

ggsave("Output/251205_siltux_volcano.pdf", width=4, height = 4)
#write.csv(c.s.results,"Output/siltux_limma.csv")
```

### Tocilizumab Limma

```{r}
num.df <- normed[21:40]
#rownames(num.df) <-t.protein_data[,1]
#select numeric
#num.df <- as.data.frame(num.df[,2:ncol(num.df)])
#head(num.df)


#Removing 0's because they don't log2
num.df[num.df==0] <- NA

#make the data a matrix of values
mat <- data.matrix(num.df)
#Limma expects a log2 value, so converting that if it is not already
l2m <- log2(mat)

#Extracting conditions
patient <- factor(meta$Patient)
treat <- factor(meta$Condition, level=c('TB','TT'))

design <- model.matrix(~patient+treat)

fit <- lmFit(l2m, design)
fits <- eBayes(fit)

#contrast.matrix <- makeContrasts(
  #Siltux_Treatment_vs_Baseline = ST - SB,
  #levels = design
#)

t.results <- topTable(fits, coef = "treatTT", adjust="BH", n=Inf)

#t.results <- topTable(fits, coef = "Tocil_Treatment_vs_Baselin")

c.t.results <- t.results
c.t.results["-log10_adj_p-val"] <- -log10(c.t.results[5])

#write.csv(c.t.results, "250226 Tocil treatment vs baseline PAIRED TIC normalized imputed.csv", row.names = TRUE)

ggplot(c.t.results, aes(x=logFC, y=`-log10_adj_p-val`))+
  geom_point() +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "red") +
  theme_minimal() +
  labs(x="Log2(Fold Change)", y="-Log10(p-value)", title = "Tocil treatment vs baseline PAIRED, TIC normalized, imputed boundaries")

c.t.results <- c.t.results %>% mutate(gene = sub(".*\\|(.*?)_.*", "\\1", row.names(c.t.results)))

sig_t_results <- c.t.results[which(c.t.results$adj.P.Val<0.05),]

```

```{r}
ggplot(c.t.results, aes(x = logFC, y = `-log10_adj_p-val`)) + 
  geom_point(color = "grey80") + 
  theme_cowplot() + 
  geom_point(data = sig_t_results, aes(x = logFC, y = `-log10_adj_p-val`), color = "black") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black") +
  geom_text_repel(
    data = sig_t_results,
    aes(x = logFC, y = `-log10_adj_p-val`, label = gene),
    color = "black",
    size = 3,
    box.padding = 0.3,
    segment.color = "black",
    max.overlaps = 25,
    fill = "white",     # gives white background bubble
    min.segment.length = 0,
    bg.color = "white", # for ggrepel ≥0.9.3, use bg.color/bg.r
    bg.r = 0.15         # rounded corners
  ) +
  
  

  
  scale_x_continuous(limits = c(-4, 4), breaks = seq(-4, 4, 1)) + 
  scale_y_continuous(limits = c(0, 4.5), breaks = seq(0, 4.5, 0.5)) + 
  scale_y_break(c(2.2,3.4)) +
  
  labs(x="Tocilizumab log2(treatment/baseline)", y="-log10(adj. P-value)")
ggsave("Output/251205_tocil_volcano.pdf", width=4, height = 4)

#write.csv(c.t.results,"Output/tocil_limma.csv")

```

## Statistical propogation for bar plots

From Wu, et al. 2024 (2023?)

Converting 95% CI to standard error:

$$SE_{FC} = \frac{FC_{max} - FC_{min}}{\ 3.92}$$

Converting standard error to standard deviation:

$$SD_{FC} = \frac{SE_{FC}}{\sqrt{\frac{1}{n_{EV}}+\frac{1}{n_{Plasma}}}}$$

Taking into account Log2 transformation on standard deviation:

$$SD_{Log_{2}FC} = \frac{1}{ln(2)}*\frac{SD_{FC}}{FC}$$

-   Requires identified significant proteins from Limma
-   Requires all_proteins_ratios

### Preparing ratios from normalized data

```{r}
df <- normed

Siltux_baseline <- select(df, starts_with(Siltux_baseline_sequence))
Siltux_treat <- select(df, starts_with(Siltux_treat_sequence))
Tocil_baseline <- select(df, starts_with(Tocil_baseline_sequence))
Tocil_treat <- select(df, starts_with(Tocil_treat_sequence))
Siltux_ratio <- Siltux_treat/Siltux_baseline
Tocil_ratio <- Tocil_treat/Tocil_baseline

all_proteins_ratios <- cbind(Siltux_ratio, Tocil_ratio)
```

#### Sorting out what's significant

```{r}
sig_s_list <- sig_s_results$gene
sig_t_list <- sig_t_results$gene
sig_s_only <- setdiff(sig_s_list, sig_t_list)
sig_t_only <- setdiff(sig_t_list, sig_s_list)
sig_both <- intersect(sig_s_list, sig_t_list)
sig_all <- union(sig_s_list, sig_t_list)

write.csv(sig_all, "sig_all.csv")
```

```{r}
CI_SEwidth <- (2*1.96)
n_treat = 10
n_control = 10
denom <- 1/(sqrt((1/n_treat)+(1/n_control)))

df <- all_proteins_ratios %>%
  mutate(gene = sub(".*\\|(.*?)_.*", "\\1", row.names(df)))%>%
  mutate("siltux_minFC" = apply(select(.,1:10),1, min, na.rm=TRUE),
         "siltux_maxFC" = apply(select(.,1:10),1, max, na.rm=TRUE),
         "siltux_mean" = rowMeans(select(.,1:10), na.rm=TRUE),
         "siltux_SE" = ((siltux_maxFC-siltux_minFC)/CI_SEwidth),
         "siltux_SD" = siltux_SE/denom,
         "siltux_mean_log2" = log2(siltux_mean),
         "siltux_SD_log2" = 1/log(2)*siltux_SD/siltux_mean) %>%
  mutate("tocil_minFC" = apply(select(.,11:20),1, min, na.rm=TRUE),
         "tocil_maxFC" = apply(select(.,11:20),1, max, na.rm=TRUE),
         "tocil_mean" = rowMeans(select(.,11:20), na.rm=TRUE),
         "tocil_SE" = ((tocil_maxFC-tocil_minFC)/CI_SEwidth),
         "tocil_SD" = tocil_SE/denom,
         "tocil_mean_log2" = log2(tocil_mean),
         "tocil_SD_log2" = 1/log(2)*tocil_SD/tocil_mean)

# Tocil only 
df_tocil_only <- df %>% filter(gene %in% sig_t_only)
df_siltux_only <- df %>% filter(gene %in% sig_s_only)
df_both <- df %>% filter(gene %in% sig_both)
df_sig_all <- df %>% filter(gene %in% sig_all)

write.csv(df_tocil_only, "Output/250416_tocil_only.csv")
write.csv(df_siltux_only, "Output/250416_siltux_only.csv")
write.csv(df_both, "Output/250416_both.csv")
write.csv(df_sig_all, "Output/sig_all.csv")
```

## Heatmaps

```{r}
df <- normed 
#df <- df %>%
  #mutate(gene = sub(".*\\|(.*?)_.*", "\\1", row.names(df))) %>%
  #filter(gene %in% sig_all)
  

txt <- "TBK1,SAA1,SQSTM1,PSMD6,PSMD13,PSMC1,PSMD2,PSMD8,IKBKB,IRAK4,PSMC5,PSMD12,PSMC2,PSMD14,RBX1,PSMC3,PSMB5,PSMC4,UBB,NFKB1,PSMD1,PSMC6,GSDMD,PSMB6,MAP3K7,PSMA5,PSMD11,PSMB4,PSMB3,PSMA4,PSMB7,PSMA7,PELI1,PSMA2,PSMB2,PSMA1,PSMA3,PSMA6,USP14"

selected <- strsplit(txt, ",")[[1]]

df <- normed_gn %>%
  filter(GeneName %in% selected)
  
row.names(df) <- df$GeneName
colnames(df) <- substr(colnames(df), 1, 3)

df_siltux <- df[4:23]
df_tocil <- df[24:43]
df_siltux_scale <- t(scale(t(df_siltux)))
df_tocil_scale <- t(scale(t(df_tocil)))

pre_treatment <- c(Siltux_baseline_sequence, Tocil_baseline_sequence)
treatment_status_siltux <- ifelse(colnames(df_siltux) %in% pre_treatment, "Pre", "Post") %>%
  factor(., levels=c("Pre", "Post"))
treatment_status_tocil <- ifelse(colnames(df_tocil) %in% pre_treatment, "Pre", "Post")%>%
  factor(., levels=c("Pre", "Post"))

treatment_annot_siltux <- HeatmapAnnotation(
  Treatment = treatment_status_siltux,
  col = list(Treatment = c("Pre" = "#FFBE45", "Post" = "#DD4AC0"))
)

treatment_annot_tocil <-HeatmapAnnotation(
  Treatment = treatment_status_tocil,
  col = list(Treatment = c("Pre" = "#FFBE45", "Post" = "#DD4AC0"))
)
```

```{r}
pdf("Output/251152_heatmap_IL1_family_siltux.pdf", width=10, height=15)


Heatmap(df_siltux_scale,
        cluster_rows = TRUE,
        cluster_columns = FALSE,
        show_row_names = TRUE,
        show_column_names = TRUE,
        column_names_side = "top",
        color = colorRampPalette(c("blue", "white", "red"))(100),
        top_annotation = treatment_annot_siltux)
dev.off()

pdf("Output/251152_heatmap_IL1_family_tocil.pdf", width=10, height=15)

Heatmap(df_tocil_scale,
        cluster_rows = TRUE,
        cluster_columns = FALSE,
        show_row_names = TRUE,
        show_column_names = TRUE,
        column_names_side = "top",
        color = colorRampPalette(c("blue", "white", "red"))(100),
        top_annotation = treatment_annot_tocil)
dev.off()

```

## X-Y Limma results

merge Limma datasets

```{r}

c.t.results <- c.t.results %>%
  mutate(ID = row.names(c.t.results))
merge <- left_join(c.t.results, c.s.results, by="gene",suffix=c(".t",".s")) %>%
  mutate(UniProtID = str_extract(ID, "(?<=\\|)[A-Z0-9]+"))

merge_sig_t <- merge %>%
  filter(adj.P.Val.t <0.05)
merge_sig_s <- merge %>%
  filter(adj.P.Val.s <0.05)
merge_sig_both <- merge %>%
  filter(adj.P.Val.t < 0.05 & adj.P.Val.s <0.05)
merge_sig_neither <- merge %>%
  filter(adj.P.Val.t > 0.05 & adj.P.Val.s >0.05)
merge_sig_either <- merge %>%
  filter(adj.P.Val.t < 0.05 | adj.P.Val.s < 0.05)

treatment_correlation <- cor(merge$logFC.t, merge$logFC.s, method="spearman")
sig_correlation <- cor(merge_sig_either$logFC.t, merge_sig_either$logFC.s)

#write.csv(merge,"250731_results.csv")
```

```{r}
ggplot(merge_sig_neither, aes(x=logFC.t, y=logFC.s))+
  geom_point(color="grey65",
             size=2,alpha=0.5,stroke=0)+
  geom_vline(xintercept=0,linetype="dashed", color="black")+
  geom_hline(yintercept=0,linetype="dashed", color="black")+
  theme_cowplot()+
  geom_point(data=merge_sig_t, aes(x=logFC.t, y=logFC.s),
             color = "#1592ed", size=2,stroke=0)+
   geom_point(data=merge_sig_s, aes(x=logFC.t, y=logFC.s),
             color = "#d72e19", size=2,stroke=0)+
  geom_point(data=merge_sig_both,  aes(x=logFC.t, y=logFC.s),
             color = "#6416df", size=2,stroke=0)+
  
  geom_text_repel(data=merge_sig_either, aes(x=logFC.t, y=logFC.s,
                                        label = gene),
    color = "black",
    size = 3,
    fontface = "bold",
    box.padding = 0.5,
    segment.color = "black",
    max.overlaps = 25,
    fill = "white",     # gives white background bubble
    min.segment.length = 0,
    bg.color = "white", # for ggrepel ≥0.9.3, use bg.color/bg.r
    bg.r = 0.15         # rounded corners
  ) + 
  
  

 
  labs(x="Tocilizumab log2(treatment/baseline)", y="Siltuximab log2(treatment/baseline)")
  
ggsave("251024_XY.pdf", width=5,height=5)
```

## patient by patient correlation

Which patients across cohorts are the same individuals: - T01/S01 - T03/S13 - T05/S05 - T07/S09 - T09/S19

### Start by filtering dataframe to these 10 samples

```{r}
pts_baseline <- list("T01", "S01", "T03", "S13", "T05", "S05", "T07", "S09", "T09", "S19")

pts_treatment <- list("T02","S02","T04","S14","T06","S06","T08","S10","T10","S20")

df_pts_baseline <- normed %>%
  select(matches(paste(pts_baseline, collapse = "|"))) %>%
  na.omit()

df_pts_treatment <- normed %>%
  select(matches(paste(pts_treatment, collapse = "|"))) %>%
  na.omit()

df_all <- normed %>%
  na.omit()
```

Next, compute each pairwise correlation.

```{r}
corrs_baseline <- cor(df_pts_baseline, method = "spearman")
corrs_treatment <- cor(df_pts_treatment)
corrs_all <- cor(df_all, method = "spearman")
```

Once again, an unsupervised clustering heatmap

```{r}
library(ComplexHeatmap)
library(circlize)
library(grid)
i=1
j=1

pdf("Output/250822_pts_baseline.pdf", width = 12, height = 12)

col_fun <- colorRamp2(c(0.8, 0.9, 1), c("blue", "white", "red"))

Heatmap(
  corrs_baseline,
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  show_row_names = TRUE,
  show_column_names = TRUE,
  column_names_side = "top",
  col = col_fun,
  cell_fun = function(j, i, x, y, width, height, fill) {
    grid.text(
      round(corrs_baseline[i, j], 3),  # value rounded to 3 decimals
      x, y,
      gp = gpar(fontsize = 15)          # adjust text size as needed
    )
  }
)

dev.off()
```

```{r}

df_pt1_norm <- as.data.frame(cbind(normed$T09_39, normed$S19_55))  %>%
  na.omit()

plot(df_pt1_norm$V1, df_pt1_norm$V2)
cor <- cor(as.numeric(df_pt1_norm$V1), as.numeric(df_pt1_norm$V2))
```

```{r}
ggplot(df_pt1_norm, aes(x = log10(V1), y = log10(V2))) +
  geom_point(alpha = 0.6, color = "black") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "blue") +
   theme_cowplot() +
  labs(x = "2016 baseline, log10 signal", y = "2017 baseline, log10 signal", title = "Same patient, blood drawn a year apart")

ggsave("ptT09-S19.pdf", height=5,width=5)
```

### Correlation plots, EV markers
```{r}
normed <- read.csv("Output/251125_normed_proteins.csv")
pairs <- read.csv("Input/pairs.csv")
Enrichment <- read.csv("Input/EnrichmentMarkersV2.csv")
Depletion <- read.csv("Input/DepletionMarkers.csv")

if (!dir.exists("Output/correlation_plots")) dir.create("Output/correlation_plots", recursive = TRUE)

plot_list <- list()

for (i in seq_len(nrow(pairs))) {

  tocil_code  <- pairs$Tocil[i]
  siltux_code <- pairs$Siltux[i]

  tocil_col  <- grep(paste0("^", tocil_code, "_"), colnames(normed), value = TRUE)
  siltux_col <- grep(paste0("^", siltux_code, "_"), colnames(normed), value = TRUE)

  if (length(tocil_col) != 1 || length(siltux_col) != 1) {
    message("Skipping pair ", tocil_code, "/", siltux_code,
            " — expected 1 match each, found Tocil:", length(tocil_col), " Siltux:", length(siltux_col))
    next
  }

  df <- normed %>%
    select(GeneName, Protein.Accession, V1 = all_of(tocil_col), V2 = all_of(siltux_col)) %>%
    filter(!is.na(V1), !is.na(V2), V1 > 0, V2 > 0)

  if (nrow(df) == 0) {
    message("Skipping pair ", tocil_code, "/", siltux_code, " — no complete positive-value points")
    next
  }

  df <- df %>%
    mutate(
      is_enriched = Protein.Accession %in% Enrichment$Uniprot.accession,
      is_depleted = Protein.Accession %in% Depletion$Uniprot.accession
    )

  enriched_labels <- df %>% filter(is_enriched)
  depleted_labels <- df %>% filter(is_depleted)

  combined_labels <- bind_rows(
    enriched_labels %>% mutate(label_color = "#D72638"),
    depleted_labels %>% mutate(label_color = "#26567E")
  )

  p <- ggplot(df, aes(x = log10(V1), y = log10(V2))) +
    geom_point(alpha = 1, color = "grey80") +
    geom_point(data = enriched_labels, aes(x = log10(V1), y = log10(V2)), color = "#D72638", size = 2) +
    geom_point(data = depleted_labels, aes(x = log10(V1), y = log10(V2)), color = "#377EB8", size = 2) +
    
    scale_color_identity() +
    #geom_smooth(method = "lm", se = FALSE, color = "red") +
   geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "black") +
    geom_text_repel(data = combined_labels, aes(x = log10(V1), y = log10(V2), label = GeneName, color = label_color),
                 size = 3, fontface = "bold",
                 box.padding = 0.6,
                 point.padding = 0.4,
                 max.overlaps = 20,
                 show.legend = FALSE,
                 force = 7,
                 force_pull = 0.3,
                 min.segment.length = 0,
                 seed = 55) +
    theme_cowplot() +
    coord_equal(xlim = c(0, 10), ylim = c(0, 10)) +
    labs(x = "2016 baseline, log10 signal", y = "2017 baseline, log10 signal",
         #title = paste0(tocil_code, " / ", siltux_code))+
         title=NULL)+
    theme(plot.margin = margin(t = 10, r = 15, b = 10, l = 10))

  plot_list[[paste0(tocil_code, "_", siltux_code)]] <- p

  ggsave(paste0("Output/correlation_plots/pt", tocil_code, "-", siltux_code, ".pdf"),
         plot = p, height = 4, width = 4)
}
```


## PCA

```{r}
normed_na_omit <- normed %>%
  na.omit()

PCA <- prcomp(t(normed_na_omit))
PCA_x <- as.data.frame(PCA$x)
PCA_prots <- as.data.frame(PCA$center)
write.csv(PCA_prots,"251002_PCA.csv")

color_pal <- c(rep(c("#F9E0D9","#DA4167"),times=10),rep(c("#89A9E4","#0A369D"), times=10))

color_pal <-rep(c("#F192AA","#DA4167","#89A9E4","#0A369D"), each=10)

plot_ly(x=PCA_x$PC1,
        y=PCA_x$PC2, 
        name = row.names(PCA_x),
        marker = list(color = color_pal, size = 9)
) %>%
  layout(title="PCA",
         xaxis=list(title="PCA1"),
         yaxis=list(title="PCA2"))

ggplot(PCA_x, aes(x=PC1, y=PC2))+
  geom_point(color=color_pal,
             size=4,stroke=0)+
  #geom_smooth(method="lm", color="black",se=TRUE)+
  theme_cowplot()+
  #geom_point(data=merge_sig_t, aes(x=logFC.t, y=logFC.s),
            # color = "#D72638", size=2,alpha=0.5,stroke=0)+
  # geom_point(data=merge_sig_s, aes(x=logFC.t, y=logFC.s),
            # color = "#377EB8", size=2,alpha=0.5,stroke=0)+
  

  labs(x="PC1", y="PC2")
  
ggsave("251024_PCA.pdf", width=5,height=4)
```

#### Histogram of PC1

```{r}
prefixes <- c(Siltux_baseline_sequence, Tocil_baseline_sequence)
pattern <- paste(prefixes, collapse = "|")  
PCA_pre <- PCA_x %>%
  filter(grepl(pattern, rownames(PCA_x)))

prefixes <- c(Siltux_treat_sequence, Tocil_treat_sequence)
pattern <- paste(prefixes, collapse = "|") 
PCA_post <- PCA_x %>%
  filter(grepl(pattern, rownames(PCA_x)))

hist_pval <- var.test(PCA_post$PC1,PCA_pre$PC1)

ggplot(PCA_pre, aes(x=PC1))+
  
  geom_density(data=PCA_post, (aes(x=PC1)),
               fill="#8833FF",alpha=0.6)+
  geom_density(fill="#CFADFF", alpha=0.6)+
  theme_cowplot()

ggsave("251024_PC1.pdf",width=5,height=2)
```

## FGSEA - log2FC

-   Fetch FGSEA code
-   Download human database
-   prepare rnk file (gene symbol, ranking metric)

Notes: reactome pathway shows IL1 signaling decreased only for tocil... and only one of the reactome sets

### Initializing fgsea:

```{r}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("reactome.db")
```

```{r}
library(fgsea)
library(reactome.db)
```

### Code for fgsea (tocil, log2FC):

```{r}
# initial seed set
#my_seed <- sample.int(.Machine$integer.max, 1)
# permanent seed moving forward:
my_seed = 573953901

set.seed(my_seed)

ranks <- read.table("input/Tocil_gene_name.rnk",
                    header=FALSE, colClasses = c("character", "numeric")) 

ranks <- ranks[!duplicated(ranks$V1), ]
ranks <- na.omit(ranks)

ranks <- setNames(ranks$V2, ranks$V1)
str(ranks)

#pathways <- gmtPathways("input/Jensen_COMPARTMENTS.gmt")
pathways <- gmtPathways("input/ReactomePathways.gmt")
#pathways <- reactomePathways(names(exampleRanks))
str(head(pathways))

fgseaRes <- fgsea(pathways,ranks, minSize=5,maxSize=500)
head(fgseaRes)

fgsea_tocil <- fgseaRes %>%
  #mutate(hits = lengths(leadingEdge)) %>%
  dplyr::mutate(leadingEdge = purrr::map_chr(leadingEdge, ~ paste(.x, collapse = ",")))
  #dplyr::select(-leadingEdge)

write.csv(fgsea_tocil, "Output/fgsea_tocil_log2FC_reactome.csv")
topPathwaysUp <- fgseaRes[ES > 0][head(order(pval), n=10), pathway]
topPathwaysDown <- fgseaRes[ES < 0][head(order(pval), n=10), pathway]

topPathways <- c(topPathwaysUp, rev(topPathwaysDown))
plotGseaTable(pathways[topPathways], ranks, fgseaRes, 
              gseaParam=0.5)
```

### Fgsea, siltux

```{r}

set.seed(my_seed)

ranks <- read.table("input/Siltux_gene_name.rnk",
                    header=FALSE, colClasses = c("character", "numeric"))
ranks <- ranks[!duplicated(ranks$V1), ]
ranks <- na.omit(ranks)
ranks <- setNames(ranks$V2, ranks$V1)

str(ranks)


#pathways <- gmtPathways("input/Jensen_COMPARTMENTS.gmt")
pathways <- gmtPathways("input/ReactomePathways.gmt")
str(head(pathways))

fgseaRes <- fgsea(pathways,ranks, minSize=5,maxSize=500)
head(fgseaRes)

fgsea_siltux <- fgseaRes %>%
  #mutate(hits = lengths(leadingEdge)) %>%
  dplyr::mutate(leadingEdge = purrr::map_chr(leadingEdge, ~ paste(.x, collapse = ",")))
  #dplyr::select(-leadingEdge)


write.csv(fgsea_siltux, "Output/fgsea_siltux_log2FC_reactome.csv")
topPathwaysUp <- fgseaRes[ES > 0][head(order(pval), n=10), pathway]
topPathwaysDown <- fgseaRes[ES < 0][head(order(pval), n=10), pathway]

topPathways <- c(topPathwaysUp, rev(topPathwaysDown))
plotGseaTable(pathways[topPathways], ranks, fgseaRes, 
              gseaParam=0.5)
```

### fgsea plotting

```{r}
library(stringr)

df <- fgsea_tocil #%>%
  #filter(grepl("Interleukin", pathway, ignore.case = TRUE))

df_top5_up <- df %>%
  filter(ES >0) %>%
  #filter(padj<0.5)%>%
  arrange(desc(NES))%>%
  dplyr::slice(1:20) %>%
  arrange(desc(NES))

df_top5_down <- df %>%
  filter(ES <0) %>%
  #filter(padj<0.5)%>%
  arrange(NES)%>%
  dplyr::slice(1:20) %>%
  arrange(desc(NES))

df_top10 <- rbind(df_top5_up, df_top5_down) %>%
  mutate(label = paste(pathway),
         label_wrapped = str_wrap(label,width=100))
library(cowplot)

df_top10_down <- df %>%
  filter(ES <0) %>%
  filter(padj<0.1)%>%
  arrange(NES)%>%
  dplyr::slice(1:10) %>%
  arrange(desc(NES)) %>%
  mutate(label=pathway,
  label_wrapped = str_wrap(label,width=8))%>%
  mutate(hits = lengths(leadingEdge))




color_limits <- c(0,0.05)



ggplot(df_top10, aes(x = reorder(label_wrapped, NES), y = NES, fill = padj)) +
  theme_cowplot()+
  geom_col() +
  coord_flip() + 
  scale_fill_gradient(low = "red", high="blue", limits = color_limits, name="Adjusted p-value")+
  labs(
    #title = "Top 10 regulated pathways from H2F treatment",
    x = "Pathway",
    y = "Normalized Enrichment Score (NES)"
  ) +

  theme(
    axis.text.y = element_text(size = 11, lineheight = 1),  # more line spacing
    plot.margin = margin(t = 10, r = 30, b = 10, l = 100)     # more room on the left
  )

ggsave("251002_tocil_fgsea.svg", width=12, height=10)
```

### Heatmap of interleukin 1 signaling

```{r}
#IL-1 leading edge from fgsea
IL1 <- fgsea_tocil %>%
  filter(pathway == "Interleukin-1 family signaling") %>%
  pull(leadingEdge) %>%
  strsplit(",") %>%
  unlist()

```

```{r}
df <- normed 
df <- df %>%
  mutate(gene = sub(".*\\|(.*?)_.*", "\\1", row.names(df))) %>%
  filter(gene %in% IL1)

row.names(df) <- df$gene
colnames(df) <- substr(colnames(df), 1, 3)

df_siltux <- df[1:20]
df_tocil <- df[21:40]
df_siltux_scale <- t(scale(t(df_siltux)))
df_tocil_scale <- t(scale(t(df_tocil)))

pre_treatment <- c(Siltux_baseline_sequence, Tocil_baseline_sequence)
treatment_status_siltux <- ifelse(colnames(df_siltux) %in% pre_treatment, "Pre", "Post") %>%
  factor(., levels=c("Pre", "Post"))
treatment_status_tocil <- ifelse(colnames(df_tocil) %in% pre_treatment, "Pre", "Post")%>%
  factor(., levels=c("Pre", "Post"))

treatment_annot_siltux <- HeatmapAnnotation(
  Treatment = treatment_status_siltux,
  col = list(Treatment = c("Pre" = "#FFBE45", "Post" = "#DD4AC0"))
)

treatment_annot_tocil <-HeatmapAnnotation(
  Treatment = treatment_status_tocil,
  col = list(Treatment = c("Pre" = "#FFBE45", "Post" = "#DD4AC0"))
)
```

```{r}

pdf("Output/251111_heatmap_siltux_IL1.pdf", width=10, height=15)


Heatmap(df_siltux_scale,
        cluster_rows = TRUE,
        cluster_columns = TRUE,
        show_row_names = TRUE,
        show_column_names = TRUE,
        column_names_side = "top",
        color = colorRampPalette(c("blue", "white", "red"))(100),
        top_annotation = treatment_annot_siltux)
dev.off()

pdf("Output/251111_heatmap_tocil_IL1.pdf", width=10, height=15)

Heatmap(df_tocil_scale,
        cluster_rows = TRUE,
        cluster_columns = TRUE,
        show_row_names = TRUE,
        show_column_names = TRUE,
        column_names_side = "top",
        color = colorRampPalette(c("blue", "white", "red"))(100),
        top_annotation = treatment_annot_tocil)
dev.off()
```

### Merging the sets

```{r}
#merge <- left_join(fgsea_tocil, fgsea_siltux, by="pathway",suffix=c(".t",".s"))

library(readxl)

merge <- read_xlsx("Output/fgsea_log2FC_reactome.xlsx", 
                   sheet = 1, skip =1,
                   range = cell_limits(c(2, 1), c(NA, 17)))

merge_sig_t <- merge %>%
  filter(padj.t <0.05)
merge_sig_s <- merge %>%
  filter(padj.s <0.05)
merge_sig_both <- merge %>%
  filter(padj.t < 0.05 & padj.s <0.05)
merge_sig_neither <- merge %>%
  filter(padj.t > 0.05 & padj.s >0.05)
merge_sig_either <- merge %>%
  filter(padj.t < 0.05 | padj.s < 0.05)

df_merged <- as.data.frame(merge)

#df_merged %>%
  #select(-where(is.list)) %>%
  #write.csv("Output/fgsea_log2FC_reactome.csv", row.names = FALSE, na = "")
```

Filter the merged set of pathways to top 5

```{r}
quad_1_t <- merge_sig_either %>%
  filter(NES.t < 0 & NES.s >0) %>%
  arrange(desc(abs(NES.t))) %>%
  dplyr::slice(1:5)
quad_1_s <- merge_sig_either %>%
  filter(NES.t < 0 & NES.s >0) %>%
  arrange(desc(abs(NES.s))) %>%
  dplyr::slice(1:5)

quad_2_t <- merge_sig_either %>%
  filter(NES.t > 0 & NES.s >0) %>%
  arrange(desc(abs(NES.t))) %>%
  dplyr::slice(1:5)
quad_2_s <- merge_sig_either %>%
  filter(NES.t > 0 & NES.s >0) %>%
  arrange(desc(abs(NES.s))) %>%
  dplyr::slice(1:5)

quad_3_t <- merge_sig_either %>%
  filter(NES.t > 0 & NES.s < 0) %>%
  arrange(desc(abs(NES.t))) %>%
  dplyr::slice(1:5)
quad_3_s <- merge_sig_either %>%
  filter(NES.t > 0 & NES.s < 0) %>%
  arrange(desc(abs(NES.s))) %>%
  dplyr::slice(1:5)

quad_4_t <- merge_sig_either %>%
  filter(NES.t < 0 & NES.s < 0) %>%
  arrange(desc(abs(NES.t))) %>%
  dplyr::slice(1:5)
quad_4_s <- merge_sig_either %>%
  filter(NES.t < 0 & NES.s < 0) %>%
  arrange(desc(abs(NES.s))) %>%
  dplyr::slice(1:5)

addtl <- merge_sig_either %>%
  filter(pathway == "Common Pathway of Fibrin Clot Formation")

labels <- list(
  quad_1_t, quad_1_s, quad_2_t, quad_2_s,
  quad_3_t, quad_3_s, quad_4_t, quad_4_s, addtl
) |>
  lapply(as.data.frame) |>
  dplyr::bind_rows() |>
  dplyr::select(where(~ !is.list(.))) |>
  dplyr::distinct(pathway, .keep_all = TRUE)
labels <- labels %>%
  mutate(label_wrapped = str_wrap(pathway, width = 20))  
```

```{r}
ggplot(merge_sig_neither, aes(x=NES.t, y=NES.s))+
  geom_point(color="grey65",
             size=2,alpha=0.5,stroke=0)+
  
  
  #geom_smooth(data=df_filtered, aes(x=X231_log2FC, y=NGF_log2FC),
            # color = "#D72638",method="lm", se=FALSE)+
  theme_cowplot()+
  geom_point(data=merge_sig_t, aes(x=NES.t, y=NES.s),
             color = "#377EB8", size=2,stroke=0)+
   geom_point(data=merge_sig_s, aes(x=NES.t, y=NES.s),
             color = "#D72638", size=2,stroke=0)+
  geom_point(data=merge_sig_both,  aes(x=NES.t, y=NES.s),
             color = "#B870DF", size=2,stroke=0)+
  
  geom_text_repel(
    data = labels,
    aes(x = NES.t, y = NES.s, label = label_wrapped),
    color = "black",
    size = 3,
    fontface = "bold",
    box.padding = 0.5,
    segment.color = "black",
    max.overlaps = 25,
    fill = "white",     # gives white background bubble
    min.segment.length = 0,
    bg.color = "white", # for ggrepel ≥0.9.3, use bg.color/bg.r
    bg.r = 0.15         # rounded corners
  ) +
  
  geom_vline(xintercept=0,linetype="dashed", color="black")+
  geom_hline(yintercept=0,linetype="dashed", color="black")+
      
 
  labs(x="Tocilizumab NES", y="Siltuximab NES")+
  theme(plot.margin = margin(t = 10, r = 10, b = 20, l = 20))#+
  #scale_x_continuous(limits = c(-2., 3), breaks = seq(-3, 3, 1)) +
  #scale_y_continuous(limits = c(-3, 3), breaks = seq(-3, 3, 1))
  
ggsave("260128_fgsea_log2FC_reactome.pdf", width=8,height=8)
```

### Discordance metric

```{r}
merge_disc <- merge %>%
  mutate(discordance = NES.t-NES.s) %>%
  arrange(discordance) %>%
  mutate(rank = c(1:nrow(merge)))

merge_disc_top5 <- merge_disc %>%
  arrange(desc(discordance)) %>%
  dplyr::slice(1:5)

merge_disc_bottom5 <- merge_disc %>%
  arrange(discordance) %>%
  dplyr::slice(1:5)

merge_middle <- merge_disc %>%
  arrange(abs(discordance)) %>%
  dplyr::slice(1:5)

merge_10 <- rbind(merge_disc_top5, merge_disc_bottom5, merge_middle) %>%
  mutate(label_wrapped = str_wrap(pathway, width = 20))  

ggplot(merge_disc, aes(x=rank, y=discordance)) +
  geom_point() +
  theme_cowplot() + 
  geom_text_repel(
    data = merge_10,
    aes(x = rank, y = discordance, label = label_wrapped),
    color = "black",
    size = 3,
    fontface = "bold",
    box.padding = 0.5,
    segment.color = "black",
    max.overlaps = 25,
    fill = "white",     # gives white background bubble
    min.segment.length = 0,
    bg.color = "white", # for ggrepel ≥0.9.3, use bg.color/bg.r
    bg.r = 0.15         # rounded corners
  ) +
  labs(x="Rank", y="Discordance metric (NES.t - NES.s)") + 
  geom_hline(yintercept=0, linetype = "dashed", color = "black")

ggsave("Output/NES discordance.pdf", width=5, height=5)
```

### Filter to pathways, sum for plotting

-   Neddylation

```{r}
pathway <- "H2AC4,H2BC26,H2BC21,H4C1,H2BC14,TNRC6A,H3C1,ELANE"

pathway <- strsplit(pathway, ",")[[1]]

df_pathway <- normed_gn %>%
  filter(GeneName %in% pathway)

df_sums <- df_pathway %>%
  summarise(across(where(is.numeric), ~ sum(.x, na.rm = TRUE)))

write.csv(df_sums,"Output/251204_clot formation.csv")
```

## Correlational analyses

### Siltux

```{r}
library(dplyr)
df <- normed
siltux <- df["Siltux_filtered",]

df <- t(df[,1:20])

siltux <- t(siltux[,1:20])
i=1

corrs <- data.frame()
for(i in 1:ncol(df)){
  name <- colnames(df)[i]

  corr_test <- cor.test(siltux, df[,i], method="pearson")
  corr <- corr_test$estimate
  p_val <- corr_test$p.value
  corrs_add <- data.frame(name, corr, p_val)
  corrs <- rbind(corrs, corrs_add)
}

corrs$adj_p_val <- p.adjust(corrs$p_val, method="BH")
corrs_siltux <- corrs
corrs_siltux <- left_join(
  corrs_siltux,
  all_peptides %>% distinct(Protein, .keep_all = TRUE),
  by = c("name" = "Protein")
)
uniprot <- read.csv("Input/Uniprot_to_gene_name.csv")
corrs_siltux_GN <- left_join(corrs_siltux,uniprot,
                            by=c("Protein.Accession"="Entry"))%>%
  dplyr::select(1:4,59)

write.csv(corrs_siltux_GN, "251024_siltux_corrs_GN_pearson.csv")
```

### Tocil

```{r}
df <- normed
tocil <- df["Tocil_filtered",]

df <- t(df[,21:40])

tocil <- t(tocil[,21:40])
i=1

corrs <- data.frame()
for(i in 1:ncol(df)){
  name <- colnames(df)[i]

  corr_test <- cor.test(tocil, df[,i], method="pearson")
  corr <- corr_test$estimate
  p_val <- corr_test$p.value
  corrs_add <- data.frame(name, corr, p_val)
  corrs <- rbind(corrs, corrs_add)
}

corrs$adj_p_val <- p.adjust(corrs$p_val, method="BH")
corrs_tocil <- corrs
corrs_tocil <- left_join(
  corrs_tocil,
  all_peptides %>% distinct(Protein, .keep_all = TRUE),
  by = c("name" = "Protein")
)

uniprot <- read.csv("Input/Uniprot_to_gene_name.csv")
corrs_tocil_GN <- left_join(corrs_tocil,uniprot,
                            by=c("Protein.Accession"="Entry"))%>%
  select(1:4,59)

write.csv(corrs_tocil_GN, "251024_tocil_corrs_GN_pearson.csv")
```

### plot

```{r}
corrs_siltux_GN <- na.omit(corrs_siltux_GN) %>%
  filter(GeneName != "#VALUE!")

sig_corr_siltux <- corrs_siltux_GN[which(corrs_siltux_GN$adj_p_val<0.05),]

ggplot(corrs_siltux_GN, aes(x = corr, y = -log10(adj_p_val))) + 
  geom_point(color = "grey80") + 
  theme_cowplot() + 
  geom_point(data = sig_corr_siltux, aes(x = corr, y = -log10(adj_p_val)), color = "black") +
  geom_text_repel(data = sig_corr_siltux, aes(x = corr, y = -log10(adj_p_val), label = GeneName), 
                  color = "black", size = 3, box.padding = 0.3) +

  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black") +
   
  #scale_x_continuous(limits = c(-2, 2), breaks = c(-2, -1.5,-1, -0.5, 0, 0.5, 1, 1.5, 2)) + 
  #scale_y_continuous(limits = c(0, 2), breaks = c(0, 0.5, 1, 1.5, 2, 2.5)) + 
  labs(x="Spearman rho", y="-log10(adj. P-value)")
ggsave("Output/251024_siltux_corr_R.pdf", width=5, height = 6)
```

```{r}
corrs_tocil_GN <- na.omit(corrs_tocil_GN) %>%
  filter(GeneName != "#VALUE!")

sig_corr_tocil <- corrs_tocil_GN[which(corrs_tocil_GN$adj_p_val<0.05),]

ggplot(corrs_tocil_GN, aes(x = corr, y = -log10(adj_p_val))) + 
  geom_point(color = "grey80") + 
  theme_cowplot() + 
  geom_point(data = sig_corr_tocil, aes(x = corr, y = -log10(adj_p_val)), color = "black") +
  geom_text_repel(data = sig_corr_tocil, aes(x = corr, y = -log10(adj_p_val), label = GeneName), 
                  color = "black", size = 3, box.padding = 0.3) +

  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black") +
   
  #scale_x_continuous(limits = c(-2, 2), breaks = c(-2, -1.5,-1, -0.5, 0, 0.5, 1, 1.5, 2)) + 
  #scale_y_continuous(limits = c(0, 2), breaks = c(0, 0.5, 1, 1.5, 2, 2.5)) + 
  labs(x="Spearman rho", y="-log10(adj. P-value)")
ggsave("Output/251024_tocil_corr_rho.pdf", width=5, height = 6)
```

### FGSEA - correlation


### Initializing fgsea:

```{r}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("reactome.db")
```

```{r}
library(fgsea)
library(reactome.db)
```

### Code for fgsea (tocil, corr):

```{r}
my_seed = 573953901
set.seed(my_seed)

ranks <- read.table("input/tocil_R.rnk",
                    header=FALSE, colClasses = c("character", "numeric")) 

ranks <- ranks[!duplicated(ranks$V1), ]
ranks <- na.omit(ranks)

ranks <- setNames(ranks$V2, ranks$V1)
str(ranks)


#pathways <- gmtPathways("input/Jensen_COMPARTMENTS.gmt")
pathways <- gmtPathways("input/ReactomePathways.gmt")

str(head(pathways))

fgseaRes <- fgsea(pathways,ranks, minSize=5,maxSize=500)
head(fgseaRes)

fgsea_tocil <- fgseaRes %>%
  #mutate(hits = lengths(leadingEdge)) %>%
  dplyr::select(-leadingEdge)

write.csv(fgsea_tocil, "Output/fgsea_all_tocil_R_reactome.csv")
topPathwaysUp <- fgseaRes[ES > 0][head(order(pval), n=10), pathway]
topPathwaysDown <- fgseaRes[ES < 0][head(order(pval), n=10), pathway]

topPathways <- c(topPathwaysUp, rev(topPathwaysDown))
plotGseaTable(pathways[topPathways], ranks, fgseaRes, 
              gseaParam=0.5)
```

### Fgsea, siltux

```{r}
my_seed = 573953901
set.seed(my_seed)

ranks <- read.table("input/siltux_R.rnk",
                    header=FALSE, colClasses = c("character", "numeric"))
ranks <- ranks[!duplicated(ranks$V1), ]
ranks <- na.omit(ranks)
ranks <- setNames(ranks$V2, ranks$V1)

str(ranks)

#pathways <- gmtPathways("input/Jensen_COMPARTMENTS.gmt")
pathways <- gmtPathways("input/ReactomePathways.gmt")
str(head(pathways))

fgseaRes <- fgsea(pathways,ranks, minSize=5,maxSize=500)
head(fgseaRes)

fgsea_siltux <- fgseaRes%>%
  #mutate(hits = lengths(leadingEdge))%>%
  dplyr::select(-leadingEdge)

write.csv(fgsea_siltux, "Output/fgsea_all_siltux_R_reactome.csv")
topPathwaysUp <- fgseaRes[ES > 0][head(order(pval), n=10), pathway]
topPathwaysDown <- fgseaRes[ES < 0][head(order(pval), n=10), pathway]

topPathways <- c(topPathwaysUp, rev(topPathwaysDown))
plotGseaTable(pathways[topPathways], ranks, fgseaRes, 
              gseaParam=0.5)
```

### plotting

```{r}
library(stringr)

df <- fgsea_tocil #%>%
  #filter(grepl("Interleukin", pathway, ignore.case = TRUE))

df_top5_up <- df %>%
  filter(ES >0) %>%
  #filter(padj<0.5)%>%
  arrange(desc(NES))%>%
  dplyr::slice(1:20) %>%
  arrange(desc(NES))

df_top5_down <- df %>%
  filter(ES <0) %>%
  #filter(padj<0.5)%>%
  arrange(NES)%>%
  dplyr::slice(1:20) %>%
  arrange(desc(NES))

df_top10 <- rbind(df_top5_up, df_top5_down) %>%
  mutate(label = paste(pathway),
         label_wrapped = str_wrap(label,width=100))
library(cowplot)

df_top10_down <- df %>%
  filter(ES <0) %>%
  filter(padj<0.1)%>%
  arrange(NES)%>%
  dplyr::slice(1:10) %>%
  arrange(desc(NES)) %>%
  mutate(label=pathway,
  label_wrapped = str_wrap(label,width=8))




color_limits <- c(0,0.05)



ggplot(df_top10, aes(x = reorder(label_wrapped, NES), y = NES, fill = padj)) +
  theme_cowplot()+
  geom_col() +
  coord_flip() + 
  scale_fill_gradient(low = "red", high="blue", limits = color_limits, name="Adjusted p-value")+
  labs(
    #title = "Top 10 regulated pathways from H2F treatment",
    x = "Pathway",
    y = "Normalized Enrichment Score (NES)"
  ) +

  theme(
    axis.text.y = element_text(size = 11, lineheight = 1),  # more line spacing
    plot.margin = margin(t = 10, r = 30, b = 10, l = 100)     # more room on the left
  )

ggsave("251028_tocil_fgsea_R.svg", width=12, height=10)
```

```{r}
#merge <- left_join(fgsea_tocil, fgsea_siltux, by="pathway",suffix=c(".t",".s"))
#merge <- read.csv("Output/fgsea_R_compartment.csv")

colnames(merge) <- colnames(merge) %>%
  sub("\\.tocil$", ".t", .) %>%
  sub("\\.siltux$", ".s", .)

merge_sig_t <- merge %>%
  filter(padj.t <0.05)
merge_sig_s <- merge %>%
  filter(padj.s <0.05)
merge_sig_both <- merge %>%
  filter(padj.t < 0.05 & padj.s <0.05)
merge_sig_neither <- merge %>%
  filter(padj.t > 0.05 & padj.s >0.05)
merge_sig_either <- merge %>%
  filter(padj.t < 0.05 | padj.s < 0.05)

#df_merged <- as.data.frame(merge)

#df_merged %>%
  #select(-where(is.list)) %>%
  #write.csv("Output/fgsea_R_reactome.csv", row.names = FALSE, na = "")
```

Filter the merged set of pathways to top 5

```{r}
quad_1_t <- merge_sig_either %>%
  filter(NES.t < 0 & NES.s >0) %>%
  arrange(desc(abs(NES.t))) %>%
  dplyr::slice(1:5)
quad_1_s <- merge_sig_either %>%
  filter(NES.t < 0 & NES.s >0) %>%
  arrange(desc(abs(NES.s))) %>%
  dplyr::slice(1:5)

quad_2_t <- merge_sig_either %>%
  filter(NES.t > 0 & NES.s >0) %>%
  arrange(desc(abs(NES.t))) %>%
  dplyr::slice(1:5)
quad_2_s <- merge_sig_either %>%
  filter(NES.t > 0 & NES.s >0) %>%
  arrange(desc(abs(NES.s))) %>%
  dplyr::slice(1:5)

quad_3_t <- merge_sig_either %>%
  filter(NES.t > 0 & NES.s < 0) %>%
  arrange(desc(abs(NES.t))) %>%
  dplyr::slice(1:5)
quad_3_s <- merge_sig_either %>%
  filter(NES.t > 0 & NES.s < 0) %>%
  arrange(desc(abs(NES.s))) %>%
  dplyr::slice(1:5)

quad_4_t <- merge_sig_either %>%
  filter(NES.t < 0 & NES.s < 0) %>%
  arrange(desc(abs(NES.t))) %>%
  dplyr::slice(1:5)
quad_4_s <- merge_sig_either %>%
  filter(NES.t < 0 & NES.s < 0) %>%
  arrange(desc(abs(NES.s))) %>%
  dplyr::slice(1:5)

labels <- list(
  quad_1_t, quad_1_s, quad_2_t, quad_2_s,
  quad_3_t, quad_3_s, quad_4_t, quad_4_s
) |>
  lapply(as.data.frame) |>
  dplyr::bind_rows() |>
  dplyr::select(where(~ !is.list(.))) |>
  dplyr::distinct(pathway, .keep_all = TRUE)
labels <- labels %>%
  mutate(label_wrapped = str_wrap(pathway, width = 20))  
```

```{r}
ggplot(merge_sig_neither, aes(x=NES.t, y=NES.s))+
  geom_point(color="grey65",
             size=2,alpha=0.5,stroke=0)+
  
  
  #geom_smooth(data=df_filtered, aes(x=X231_log2FC, y=NGF_log2FC),
            # color = "#D72638",method="lm", se=FALSE)+
  theme_cowplot()+
  geom_point(data=merge_sig_t, aes(x=NES.t, y=NES.s),
             color = "#377EB8", size=2,stroke=0)+
   geom_point(data=merge_sig_s, aes(x=NES.t, y=NES.s),
             color = "#D72638", size=2,stroke=0)+
  geom_point(data=merge_sig_both,  aes(x=NES.t, y=NES.s),
             color = "#B870DF", size=2,stroke=0)+
  geom_vline(xintercept=0,linetype="dashed", color="black")+
  geom_hline(yintercept=0,linetype="dashed", color="black")+
 geom_text_repel(
    data = labels,
    aes(x = NES.t, y = NES.s, label = label_wrapped),
    color = "black",
    size = 3,
    fontface = "bold",
    box.padding = 0.5,
    segment.color = "black",
    max.overlaps = 15,
    fill = "white",     # gives white background bubble
    min.segment.length = 0,
    bg.color = "white", # for ggrepel ≥0.9.3, use bg.color/bg.r
    bg.r = 0.15         # rounded corners
  ) +
  
  
      
 
  labs(x="Tocilizumab NES", y="Siltuximab NES")+
  theme(plot.margin = margin(t = 10, r = 10, b = 20, l = 20))#+
  #scale_x_continuous(limits = c(-2., 3), breaks = seq(-3, 3, 1)) +
  #scale_y_continuous(limits = c(-3, 3), breaks = seq(-3, 3, 1))
  
ggsave("260206_fgsea_XY_log2FC_compart.pdf", width=7,height=7)
```

Merge

```{r}
fgsea_l2FC_tocil <- read.csv("Output/fgsea_tocil_log2FC_compart.csv")
fgsea_l2FC_siltux <- read.csv("Output/fgsea_siltux_log2FC_compartment.csv")


merge <- left_join(fgsea_l2FC_tocil, fgsea_l2FC_siltux, by="pathway", suffix = c(".tocil",".siltux"))

#write.csv(merge_R, "Output/fgsea_pearson_reactome.csv")

#ggplot(merge_R, aes(x=NES.tocil, y=NES.siltux))+
 # geom_point()
```


## Concordance analysis

```{r}
# toci and silt: limma topTable output, one row per protein group,
# with columns logFC and adj.P.Val, rownames = protein group ID

limma_results <- read.csv("Output/limma_results.csv")

#both <- inner_join(
#  toci %>% tibble::rownames_to_column("protein") %>%
#           select(protein, logFC_T = logFC, adjP_T = adj.P.Val),
#  silt %>% tibble::rownames_to_column("protein") %>%
#           select(protein, logFC_S = logFC, adjP_S = adj.P.Val),
#  by = "protein"
#) %>%
#  filter(!protein %in% c("Siltux", "Tocil"))   # drop the drugs

both <- limma_results %>%
  filter(!gene %in% c("Siltux_filtered", "Tocil_filtered"))

sig_union <- both %>% filter(adj.P.Val.t < 0.05 | adj.P.Val.s < 0.05)
sig_toci  <- both %>% filter(adj.P.Val.t < 0.05)
sig_siltux <- both %>% filter(adj.P.Val.s < 0.05)

concord <- function(d) {
  agree <- sum(sign(d$logFC.t) == sign(d$logFC.s))
  n     <- nrow(d)
  bt    <- binom.test(agree, n, p = 0.5, alternative = "two.sided")
  list(agree = agree, n = n, pct = round(100 * agree / n, 1), p = bt$p.value)
}

concord(sig_union)
concord(sig_toci)
concord(sig_siltux)
```

