---
title: "IL6 Serum Biologics"
output: html_notebook
---

## Libraries

```{r}
library(cowplot)
library(ggplot2)
library(limma)
library(dplyr)
library(plotly)
library(ggrepel)
library(svglite)
library(ComplexHeatmap)
library(tidyverse)
```


## Import dataframes

```{r}
all_peptides <- read.csv("Input/250121_T1D_Serum_EV_Treatment.csv",
                         header=TRUE, 
                         na = "#N/A") %>%
  # This protein has significant peptide overlap with Siltuximab
  # It is removed here to avoid spurious results/incorrect quantitation
  # Open to other ideas
  filter(!grepl("HV145",Protein)) 



meta <- read.csv("Input/250125_meta.csv",
                 stringsAsFactors = FALSE)
```

### Roll up into proteins

```{r}
protein_data <- all_peptides %>%
  group_by(Protein) %>%
  summarise(across(7:ncol(all_peptides)-1, sum, na.rm = TRUE))
protein_data <- protein_data %>%
  mutate(across(-Protein, ~ replace(., . == 0, NA)))
```




### Data manipulation, mAb peptide-level

#### Siltuximab

- Results in "s.peptide_data", a dataframe with only siltuximab peptides in siltuximab treatment
- Writes siltux_peptides.csv for plotting

```{r}
df <- all_peptides %>%
  filter(grepl("siltux", Protein, ignore.case=TRUE))

df$Protein_peptide <- paste0(df$Protein,"_",df$Peptide.Modified.Sequence)

row.names(df) <- df$Protein_peptide

Siltux_baseline_sequence <- sprintf("S%02d",seq(1,19,by=2))
Siltux_treat_sequence <- sprintf("S%02d",seq(2,20,by=2))

Tocil_baseline_sequence <- sprintf("T%02d",seq(1,19,by=2))
Tocil_treat_sequence <- sprintf("T%02d",seq(2,20,by=2))

Siltux_baseline_peptides <- select(df, starts_with(Siltux_baseline_sequence))
Siltux_treat_peptides <- select(df, starts_with(Siltux_treat_sequence))

s.peptide_data <- cbind(Siltux_baseline_peptides, Siltux_treat_peptides)

row.names(s.peptide_data) <- row.names(df)

write.csv(s.peptide_data, "Output/siltux_peptides.csv")
```

#### Tocilizumab

- Results in "t.peptide_data", a dataframe with only siltuximab peptides in siltuximab treatment
- Writes tocilizumab_peptides.csv for plotting

```{r}
df <- all_peptides %>%
  filter(grepl("tocil", Protein, ignore.case=TRUE))

df$Protein_peptide <- paste0(df$Protein,"_",df$Peptide.Modified.Sequence)

row.names(df) <- df$Protein_peptide

#Siltux_baseline_sequence <- sprintf("S%02d",seq(1,19,by=2))
#Siltux_treat_sequence <- sprintf("S%02d",seq(2,20,by=2))

#Tocil_baseline_sequence <- sprintf("T%02d",seq(1,19,by=2))
#Tocil_treat_sequence <- sprintf("T%02d",seq(2,20,by=2))

Tocil_baseline_peptides <- select(df, starts_with(Tocil_baseline_sequence))
Tocil_treat_peptides <- select(df, starts_with(Tocil_treat_sequence))

t.peptide_data <- cbind(Tocil_baseline_peptides, Tocil_treat_peptides)

row.names(t.peptide_data) <- row.names(df)

write.csv(t.peptide_data,"Output/tocilizumab_peptides.csv")
```

### Rolling up mAbs into protein-level from a select few peptides

- Requires an input, manually-filtered set of mAb peptides
  - These were peptides belonging to the mAbs which satisfied the two following conditions:
    - Were unique to the mAb (not found in other, endogenous Igs)
      - NOTE: one peptide quantifying siltuximab (SEDTA...) overlaps with HV145 but is seemingly mostly coming from siltuximab given that it correlates with other siltuximab peptides. Consider removing this peptide from Ab quantitation
    - Significantly increased in treatment (t-tests performed in Prism)
  - Does not require running the above two sections

```{r}
quant_peptides <- read.csv("Input/250124_treatment_peptides.csv", header = FALSE)

all_peptides_siltux <- all_peptides %>%
  filter(grepl("siltux", Protein, ignore.case=TRUE))

all_peptides_siltux$Protein_peptide <- 
  paste0(all_peptides_siltux$Protein,"_", 
  all_peptides_siltux$Peptide.Modified.Sequence)

query <- quant_peptides
rownames(all_peptides_siltux) <- all_peptides_siltux$Protein_peptide

all_rownames <- rownames(all_peptides_siltux)

query <- query[[1]]
matches <- sapply(all_rownames, function(protein) {
any(sapply(query, function(name) grepl(name, protein)))
})
filtered_query <- all_peptides_siltux[matches, ]

siltux_good_peptides <- filtered_query

siltux_quant <- siltux_good_peptides %>%
  summarise(across(where(is.numeric), sum, na.rm=TRUE))

siltux_quant$Protein = "Siltux_filtered"

all_peptides_tocil <- all_peptides %>%
  filter(grepl("tocil", Protein, ignore.case=TRUE))

all_peptides_tocil$Protein_peptide <- paste0(all_peptides_tocil$Protein,"_", all_peptides_tocil$Peptide.Modified.Sequence)



query <- quant_peptides
rownames(all_peptides_tocil) <- all_peptides_tocil$Protein_peptide

all_rownames <- rownames(all_peptides_tocil)

query <- query[[1]]
matches <- sapply(all_rownames, function(protein) {
any(sapply(query, function(name) grepl(name, protein)))
})
filtered_query <- all_peptides_tocil[matches, ]

tocil_good_peptides <- filtered_query

tocil_quant <- tocil_good_peptides %>%
  summarise(across(where(is.numeric), sum, na.rm=TRUE))

tocil_quant$Protein = "Tocil_filtered"

protein_data <- rbind(protein_data, tocil_quant) %>%
  rbind(., siltux_quant)
```

### Data manipulation, protein-level, with quantitation of tocilizumab and situximab included


```{r}
all_proteins <- protein_data

row.names(all_proteins) <- all_proteins$Protein

Siltux_baseline_sequence <- sprintf("S%02d",seq(1,19,by=2))
Siltux_treat_sequence <- sprintf("S%02d",seq(2,20,by=2))

Tocil_baseline_sequence <- sprintf("T%02d",seq(1,19,by=2))
Tocil_treat_sequence <- sprintf("T%02d",seq(2,20,by=2))

Siltux_baseline <- select(all_proteins, starts_with(Siltux_baseline_sequence))
Siltux_treat <- select(all_proteins, starts_with(Siltux_treat_sequence))

s.protein_data <- cbind(Siltux_baseline, Siltux_treat)

row.names(s.protein_data) <- row.names(all_proteins)

#write.csv(s.protein_data, "250125_siltux.csv")

Tocil_baseline <- select(all_proteins, starts_with(Tocil_baseline_sequence))
Tocil_treat <- select(all_proteins, starts_with(Tocil_treat_sequence))

t.protein_data <- cbind(Tocil_baseline, Tocil_treat)

row.names(t.protein_data) <- row.names(all_proteins)

#write.csv(t.protein_data,"250121_tocil.csv")

all_samples <- cbind(s.protein_data, t.protein_data)
```


## Protein-level equalize medians

```{r}
df <- all_samples

i=1


info <- data.frame()
for(i in 1:ncol(df)){
  name <- colnames(df)[i]
  intensities <- df[,i]
  median <- median(intensities, na.rm=TRUE)
  var <- var(intensities, na.rm=TRUE)
  df_add <- data.frame(name, median, var)
  info <- rbind(info, df_add)
}

ggplot(info, aes(x = name, y = median)) +
  geom_col(fill = "steelblue") +
  theme_cowplot() +
  labs(x = "Name", y = "Median") +
  theme(axis.text.x = element_text(angle=90, hjust=0.5, vjust=0.5, size=10))

normed <- data.frame(row.names = row.names(df))
i=1

for(i in 1:ncol(df)){
  normed[,i]=df[,i]/info$median[i]
}
colnames(normed) <- colnames(df)

normed_info <- data.frame()
for(i in 1:ncol(normed)){
  name <- colnames(normed)[i]
  intensities <- normed[,i]
  median <- median(intensities, na.rm=TRUE)
  var <- var(intensities, na.rm=TRUE)
  df_add <- data.frame(name, median, var)
  normed_info <- rbind(normed_info, df_add)
}

ggplot(normed_info, aes(x = name, y = median)) +
  geom_col(fill = "steelblue") +
  theme_cowplot() +
  labs(x = "Name", y = "Median") +
  theme(axis.text.x = element_text(angle=90, hjust=0.5, vjust=0.5, size=10))
```

## Limma, volcano plots

### Siltuximab Limma
```{r}

#Select only siltuximab treatment samples
num.df <- normed[1:20]
#rownames(num.df) <-s.protein_data[,1]

#Removing 0's because they don't log2
num.df[num.df==0] <- NA

#make the data a matrix of values
mat <- data.matrix(num.df)
#Limma expects a log2 transform
l2m <- log2(mat)

#Extracting conditions from metadata file
patient <- factor(meta$Patient)
treat <- factor(meta$Condition, level=c('SB','ST'))

design <- model.matrix(~patient+treat)

fit <- lmFit(l2m, design)
fits <- eBayes(fit)

#contrast.matrix <- makeContrasts(
  #Siltux_Treatment_vs_Baseline = ST - SB,
 # levels = design
#)

s.results <- topTable(fits, coef = "treatST", adjust="BH", n=Inf)
c.s.results <- s.results
c.s.results["-log10_adj_p-val"] <- -log10(c.s.results[5])

c.s.results <- c.s.results %>% mutate(gene = sub(".*\\|(.*?)_.*", "\\1", row.names(c.s.results)))
sig_s_results <- c.s.results[which(c.s.results$adj.P.Val<0.05),]

#write.csv(c.s.results, "250219 Siltux treatment vs baseline PAIRED extra_median normalized imputed.csv", row.names = TRUE)


ggplot(c.s.results, aes(x = logFC, y = `-log10_adj_p-val`)) + 
  geom_point(color = "grey80") + 
  theme_cowplot() + 
  geom_point(data = sig_s_results, aes(x = logFC, y = `-log10_adj_p-val`), color = "black") +
  geom_text_repel(data = sig_s_results, aes(x = logFC, y = `-log10_adj_p-val`, label = gene), 
                  color = "black", size = 3, box.padding = 0.3) +

  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black") +
   
  scale_x_continuous(limits = c(-2, 2), breaks = c(-2, -1.5,-1, -0.5, 0, 0.5, 1, 1.5, 2)) + 
  scale_y_continuous(limits = c(0, 2), breaks = c(0, 0.5, 1, 1.5, 2, 2.5)) + 
  theme(axis.title = element_blank())
ggsave("Output/250227_siltux_volcano.svg", width=6, height = 8)
write.csv(c.s.results,"Output/siltux_limma.csv")
```

### Tocilizumab Limma
```{r}
num.df <- normed[21:40]
#rownames(num.df) <-t.protein_data[,1]
#select numeric
#num.df <- as.data.frame(num.df[,2:ncol(num.df)])
#head(num.df)


#Removing 0's because they don't log2
num.df[num.df==0] <- NA

#make the data a matrix of values
mat <- data.matrix(num.df)
#Limma expects a log2 value, so converting that if it is not already
l2m <- log2(mat)

#Extracting conditions
patient <- factor(meta$Patient)
treat <- factor(meta$Condition, level=c('TB','TT'))

design <- model.matrix(~patient+treat)

fit <- lmFit(l2m, design)
fits <- eBayes(fit)

#contrast.matrix <- makeContrasts(
  #Siltux_Treatment_vs_Baseline = ST - SB,
  #levels = design
#)

t.results <- topTable(fits, coef = "treatTT", adjust="BH", n=Inf)

#t.results <- topTable(fits, coef = "Tocil_Treatment_vs_Baselin")

c.t.results <- t.results
c.t.results["-log10_adj_p-val"] <- -log10(c.t.results[5])

#write.csv(c.t.results, "250226 Tocil treatment vs baseline PAIRED TIC normalized imputed.csv", row.names = TRUE)

ggplot(c.t.results, aes(x=logFC, y=`-log10_adj_p-val`))+
  geom_point() +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "red") +
  theme_minimal() +
  labs(x="Log2(Fold Change)", y="-Log10(p-value)", title = "Tocil treatment vs baseline PAIRED, TIC normalized, imputed boundaries")

c.t.results <- c.t.results %>% mutate(gene = sub(".*\\|(.*?)_.*", "\\1", row.names(c.t.results)))

sig_t_results <- c.t.results[which(c.t.results$adj.P.Val<0.05),]

ggplot(c.t.results, aes(x = logFC, y = `-log10_adj_p-val`)) + 
  geom_point(color = "grey80") + 
  theme_cowplot() + 
  geom_point(data = sig_t_results, aes(x = logFC, y = `-log10_adj_p-val`), color = "black") +
  geom_text_repel(data = sig_t_results, aes(x = logFC, y = `-log10_adj_p-val`, label = gene), 
                  color = "black", size = 3, box.padding = 0.3, max.overlaps = 15) +

  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black") +
  
  scale_x_continuous(limits = c(-2, 2), breaks = c(-2, -1.5,-1, -0.5, 0, 0.5, 1, 1.5, 2)) + 
  scale_y_continuous(limits = c(0, 2), breaks = c(0, 0.5, 1, 1.5, 2, 2.5)) + 
  theme(axis.title = element_blank())
ggsave("Output/250227_tocil_volcano.svg", width=6, height = 8)

write.csv(c.t.results,"Output/tocil_limma.csv")

```


## Statistical propogation for bar plots

From Wu, et al. 2024 (2023?)

Converting 95% CI to standard error:

  $$SE_{FC} = \frac{FC_{max} - FC_{min}}{\ 3.92}$$

Converting standard error to standard deviation:

  $$SD_{FC} = \frac{SE_{FC}}{\sqrt{\frac{1}{n_{EV}}+\frac{1}{n_{Plasma}}}}$$
  
  
Taking into account Log2 transformation on standard deviation:

  $$SD_{Log_{2}FC} = \frac{1}{ln(2)}*\frac{SD_{FC}}{FC}$$

- Requires identified significant proteins from Limma
- Requires all_proteins_ratios

### Preparing ratios from normalized data
```{r}
df <- normed

Siltux_baseline <- select(df, starts_with(Siltux_baseline_sequence))
Siltux_treat <- select(df, starts_with(Siltux_treat_sequence))
Tocil_baseline <- select(df, starts_with(Tocil_baseline_sequence))
Tocil_treat <- select(df, starts_with(Tocil_treat_sequence))
Siltux_ratio <- Siltux_treat/Siltux_baseline
Tocil_ratio <- Tocil_treat/Tocil_baseline

all_proteins_ratios <- cbind(Siltux_ratio, Tocil_ratio)
```

#### Sorting out what's significant
```{r}
sig_s_list <- sig_s_results$gene
sig_t_list <- sig_t_results$gene
sig_s_only <- setdiff(sig_s_list, sig_t_list)
sig_t_only <- setdiff(sig_t_list, sig_s_list)
sig_both <- intersect(sig_s_list, sig_t_list)
sig_all <- union(sig_s_list, sig_t_list)


```


```{r}
CI_SEwidth <- (2*1.96)
n_treat = 10
n_control = 10
denom <- 1/(sqrt((1/n_treat)+(1/n_control)))

df <- all_proteins_ratios %>%
  mutate(gene = sub(".*\\|(.*?)_.*", "\\1", row.names(df)))%>%
  mutate("siltux_minFC" = apply(select(.,1:10),1, min, na.rm=TRUE),
         "siltux_maxFC" = apply(select(.,1:10),1, max, na.rm=TRUE),
         "siltux_mean" = rowMeans(select(.,1:10), na.rm=TRUE),
         "siltux_SE" = ((siltux_maxFC-siltux_minFC)/CI_SEwidth),
         "siltux_SD" = siltux_SE/denom,
         "siltux_mean_log2" = log2(siltux_mean),
         "siltux_SD_log2" = 1/log(2)*siltux_SD/siltux_mean) %>%
  mutate("tocil_minFC" = apply(select(.,11:20),1, min, na.rm=TRUE),
         "tocil_maxFC" = apply(select(.,11:20),1, max, na.rm=TRUE),
         "tocil_mean" = rowMeans(select(.,11:20), na.rm=TRUE),
         "tocil_SE" = ((tocil_maxFC-tocil_minFC)/CI_SEwidth),
         "tocil_SD" = tocil_SE/denom,
         "tocil_mean_log2" = log2(tocil_mean),
         "tocil_SD_log2" = 1/log(2)*tocil_SD/tocil_mean)

# Tocil only 
df_tocil_only <- df %>% filter(gene %in% sig_t_only)
df_siltux_only <- df %>% filter(gene %in% sig_s_only)
df_both <- df %>% filter(gene %in% sig_both)
df_sig_all <- df %>% filter(gene %in% sig_all)

write.csv(df_tocil_only, "Output/250416_tocil_only.csv")
write.csv(df_siltux_only, "Output/250416_siltux_only.csv")
write.csv(df_both, "Output/250416_both.csv")
write.csv(df_sig_all, "Output/sig_all.csv")
```

####Put some nice ggbar plots here
```{r}


```


## Heatmaps

```{r}
df <- normed 
df <- df %>%
  mutate(gene = sub(".*\\|(.*?)_.*", "\\1", row.names(df))) %>%
  filter(gene %in% sig_all)

row.names(df) <- df$gene
colnames(df) <- substr(colnames(df), 1, 3)

df_siltux <- df[1:20]
df_tocil <- df[21:40]
df_siltux_scale <- t(scale(t(df_siltux)))
df_tocil_scale <- t(scale(t(df_tocil)))

pre_treatment <- c(Siltux_baseline_sequence, Tocil_baseline_sequence)
treatment_status_siltux <- ifelse(colnames(df_siltux) %in% pre_treatment, "Pre", "Post") %>%
  factor(., levels=c("Pre", "Post"))
treatment_status_tocil <- ifelse(colnames(df_tocil) %in% pre_treatment, "Pre", "Post")%>%
  factor(., levels=c("Pre", "Post"))

treatment_annot_siltux <- HeatmapAnnotation(
  Treatment = treatment_status_siltux,
  col = list(Treatment = c("Pre" = "#FFBE45", "Post" = "#DD4AC0"))
)

treatment_annot_tocil <-HeatmapAnnotation(
  Treatment = treatment_status_tocil,
  col = list(Treatment = c("Pre" = "#FFBE45", "Post" = "#DD4AC0"))
)
```


```{r}
pdf("Output/250416_heatmap_siltux.pdf", width=10, height=15)


Heatmap(df_siltux_scale,
        cluster_rows = TRUE,
        cluster_columns = TRUE,
        show_row_names = TRUE,
        show_column_names = TRUE,
        column_names_side = "top",
        color = colorRampPalette(c("blue", "white", "red"))(100),
        top_annotation = treatment_annot_siltux)
dev.off()

pdf("Output/250416_heatmap_tocil.pdf", width=10, height=15)

Heatmap(df_tocil_scale,
        cluster_rows = TRUE,
        cluster_columns = TRUE,
        show_row_names = TRUE,
        show_column_names = TRUE,
        column_names_side = "top",
        color = colorRampPalette(c("blue", "white", "red"))(100),
        top_annotation = treatment_annot_tocil)
dev.off()

```

