---
title: "Serum v Plasma"
output: html_notebook
editor_options: 
  markdown: 
    wrap: 72
---

todo x Fix uniprot gene name issue by joining with uniprot name
database - Correlation serum enrichment v plasma enrichment - Mag-Net
not MagNet... - I want a lighter purple color - PDF, not PNG - PCA of
samples

### Libraries

```{r}
library(cowplot)
library(ggplot2)
library(limma)
library(dplyr)
library(plotly)
library(ggrepel)
library(svglite)
library(ComplexHeatmap)
library(tidyverse)
library(ggtext)
library(ggbeeswarm)
library(ggsci)
library(cowplot)
library(dplyr)
library(tidyr)
library(stringr)
```

### Import peptides

```{r}
SVP_peptides <- read.csv("Input/SerumVPlasma_MS2.csv",
  header=TRUE, 
  na = "#N/A")%>%
  mutate(gene_name = str_extract(Protein, "(?<=\\|)[^|]+(?=_HUMAN)"))%>%
  select(gene_name, everything())

Uniprot <- read.csv("Input/Uniprot_to_gene_nameV2.csv") %>%
  select(Entry, GeneName)

Uniprot_full <- read.csv("Input/Uniprot_to_gene_nameV2.csv")

# Workaround for *values
id_cols <- c("Protein", "Protein.Name", "Protein.Description", "Protein.Accession", "gene_name", "Peptide", "Peptide.Modified.Sequence")
peptide_cols <- setdiff(names(SVP_peptides), id_cols)  # replace with your actual peptide/intensity column names

for (col in peptide_cols) {
  SVP_peptides[[col]] <- as.numeric(gsub("^\\*", "", SVP_peptides[[col]]))
}

id_cols <- c("Protein", "Protein.Name", "Protein.Description", "Protein.Accession","gene_name")
sample_cols <- setdiff(names(SVP_peptides), c(id_cols, "Peptide", "Peptide.Modified.Sequence"))

protein_data <- SVP_peptides %>%
  group_by(across(all_of(id_cols))) %>%
  summarise(across(all_of(sample_cols), sum, na.rm = TRUE), .groups = "drop")

colnames(protein_data) <- sub("_\\d+\\.Normalized\\.Area$", "", colnames(protein_data))

protein_data <- protein_data %>%
  #mutate(across(-Protein, ~ replace(., . == 0, NA))) %>%
 mutate(Plasma_EV_avg = rowMeans(select(., matches("^P\\d{2}")))) %>%
  mutate(Serum_EV_avg = rowMeans(select(., matches("^S\\d{2}")))) %>%
  mutate(Plasma_total_avg = rowMeans(select(., contains("PlasmaTotal")))) %>%
  mutate(Serum_total_avg = rowMeans(select(., contains("SerumTotal")))) %>%
  left_join(., Uniprot, by=c("Protein.Accession" = "Entry"))

group_map <- list(
  "Plasma_EV" = grep("^P\\d{2}", colnames(protein_data), value = TRUE),
  "Serum_EV"  = grep("^S\\d{2}", colnames(protein_data), value = TRUE)
)

plasma_samples <- group_map[["Plasma_EV"]]
serum_samples  <- group_map[["Serum_EV"]]

protein_data_FC <- protein_data %>%
  mutate(across(all_of(plasma_samples), ~ log2(. / Plasma_total_avg), .names = "{.col}_log2FC")) %>%
  mutate(across(all_of(serum_samples), ~ log2(. / Serum_total_avg), .names = "{.col}_log2FC"))
```

Markers for enrichment:

```{r}
Enriched <- read.csv("Input/EnrichmentMarkersV2.csv")
Depleted <- read.csv("Input/DepletionMarkers.csv")
Coagulation <- read.csv("Input/Coagulation.csv")
Platelet <- read.csv("Input/PlateletV2.csv")
Clot_formation <- read.csv("Input/reactome_clot_formation.csv")

```

### Barplot by uniprot accession

```{r}
library(ggtext)
library(ggbeeswarm)
library(ggsci)
library(cowplot)
library(dplyr)
library(tidyr)
library(stringr)

selected <- Coagulation$Uniprot.accession

group_map <- list(
  "Plasma_EV"    = grep("^P\\d{2}", colnames(protein_data), value = TRUE),
  "Serum_EV"     = grep("^S\\d{2}", colnames(protein_data), value = TRUE),
  #"Ref_EV"  = grep("^R\\d{2}_", colnames(protein_data), value = TRUE),
  "Plasma_total" = grep("PlasmaTotal", colnames(protein_data), value = TRUE),
  "Serum_total"  = grep("SerumTotal", colnames(protein_data), value = TRUE)
)

sample_cols <- unlist(group_map, use.names = FALSE)



if (!dir.exists("Output/barplots")) dir.create("Output/barplots", recursive = TRUE)

library(patchwork)

plot_list <- list()

for (i in seq_along(selected)) {

  df_plot <- protein_data %>%
    filter(Protein.Accession == selected[i])

  if (nrow(df_plot) == 0) {
    message("Skipping missing protein: ", selected[i])
    next
  }

  label_samples <- c("P03", "P04")
  
  df_long <- df_plot %>%
    select(GeneName, all_of(sample_cols)) %>%
    pivot_longer(cols = all_of(sample_cols),
                 names_to = "sample",
                 values_to = "normalized abundance") %>%
    mutate(condition = case_when(
      sample %in% group_map[["Plasma_EV"]]    ~ "Plasma_EV",
      sample %in% group_map[["Serum_EV"]]     ~ "Serum_EV",
      #sample %in% group_map[["Ref_EV"]]       ~ "Ref_EV",
      sample %in% group_map[["Plasma_total"]] ~ "Plasma_total",
      sample %in% group_map[["Serum_total"]]  ~ "Serum_total"
    )) %>%
    filter(!is.na(`normalized abundance`)) %>%
    mutate(sample_label = case_when(
    grepl(paste0("^", label_samples, collapse = "|"), sample) ~ sub("_.*", "", sample),
    TRUE ~ NA_character_
  ))

  if (nrow(df_long) == 0) {
    message("Skipping protein with no non-missing values: ", selected[i])
    next
  }

  df_long$condition <- factor(df_long$condition,
                               levels = c("Plasma_EV", "Serum_EV", "Ref_EV", "Plasma_total", "Serum_total"))

  p <- ggplot(df_long, aes(x = condition, y = `normalized abundance`, fill = condition)) +
    stat_summary(fun = mean,
                 geom = "bar",
                 color = "black",
                 width = 0.7,
                 alpha = 0.5) +
    stat_summary(fun.data = mean_sdl,
                 fun.args = list(mult = 1),
                 geom = "errorbar", width = 0.2) +
    geom_quasirandom(width = 0.2, size = 2, shape = 21, color = "black") +
   # geom_text_repel(aes(label = sample_label),
   #              color = "black", size = 3, fontface = "bold",
   #              na.rm = TRUE, box.padding = 0.4, max.overlaps = Inf) +
    scale_fill_manual(values = c(
    "Plasma_EV"    = "#9c5ae2",
    "Serum_EV"     = "#F2C438", 
    "Plasma_total" = "#377EB8",
    "Serum_total"  = "#D72638"
  )) +
    theme_cowplot() +
    labs(
      title = df_long$GeneName[[1]],
      y = "normalized abundance",
      x = NULL
    ) +
    theme(
      panel.background = element_rect(fill = "white", color = NA),
      plot.background = element_rect(fill = "white", color = NA),
      panel.grid.major.x = element_blank(),
      panel.grid.minor.x = element_blank(),
      legend.position = "none",
      axis.text.x = element_text(angle = 45, hjust = 1, size = 12)
    ) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.1)))

  plot_list[[selected[i]]] <- p
}

n_plots <- length(plot_list)
n_cols  <- 5
n_rows  <- ceiling(n_plots / n_cols)

combined <- wrap_plots(plot_list, ncol = n_cols)

ggsave("Output/barplots/Coag_MS2.pdf",
       plot = combined,
       width = n_cols * 4,
       height = n_rows * 4,
       limitsize = FALSE)
```

### Barplot by uniprot gene name

```{r}
library(ggtext)
library(ggbeeswarm)
library(ggsci)
library(cowplot)
library(dplyr)
library(tidyr)
library(stringr)

selected <- Clot_formation$gene_name

group_map <- list(
  "Plasma_EV"    = grep("^P\\d{2}_", colnames(protein_data), value = TRUE),
  "Serum_EV"     = grep("^S\\d{2}_", colnames(protein_data), value = TRUE),
  #"Ref_EV"  = grep("^R\\d{2}_", colnames(protein_data), value = TRUE),
  "Plasma_total" = grep("PlasmaTotal", colnames(protein_data), value = TRUE),
  "Serum_total"  = grep("SerumTotal", colnames(protein_data), value = TRUE)
)

sample_cols <- unlist(group_map, use.names = FALSE)

if (!dir.exists("Output/barplots")) dir.create("Output/barplots", recursive = TRUE)

library(patchwork)

plot_list <- list()

for (i in seq_along(selected)) {

  df_plot <- protein_data %>%
    filter(GeneName == selected[i])

  if (nrow(df_plot) == 0) {
    message("Skipping missing protein: ", selected[i])
    next
  }

  df_long <- df_plot %>%
    select(GeneName, all_of(sample_cols)) %>%
    pivot_longer(cols = all_of(sample_cols),
                 names_to = "sample",
                 values_to = "normalized abundance") %>%
    mutate(condition = case_when(
      sample %in% group_map[["Plasma_EV"]]    ~ "Plasma_EV",
      sample %in% group_map[["Serum_EV"]]     ~ "Serum_EV",
      #sample %in% group_map[["Ref_EV"]]       ~ "Ref_EV",
      sample %in% group_map[["Plasma_total"]] ~ "Plasma_total",
      sample %in% group_map[["Serum_total"]]  ~ "Serum_total"
    )) %>%
    filter(!is.na(`normalized abundance`))

  if (nrow(df_long) == 0) {
    message("Skipping protein with no non-missing values: ", selected[i])
    next
  }

  df_long$condition <- factor(df_long$condition,
                               levels = c("Plasma_EV", "Serum_EV", "Ref_EV", "Plasma_total", "Serum_total"))

  p <- ggplot(df_long, aes(x = condition, y = `normalized abundance`, fill = condition)) +
    stat_summary(fun = mean,
                 geom = "bar",
                 color = "black",
                 width = 0.7,
                 alpha = 0.5) +
    stat_summary(fun.data = mean_sdl,
                 fun.args = list(mult = 1),
                 geom = "errorbar", width = 0.2) +
    geom_quasirandom(width = 0.2, size = 2, shape = 21, color = "black") +
    scale_fill_viridis_d() +
    theme_cowplot() +
    labs(
      title = df_long$GeneName[[1]],
      y = "normalized abundance",
      x = NULL
    ) +
    theme(
      panel.background = element_rect(fill = "white", color = NA),
      plot.background = element_rect(fill = "white", color = NA),
      panel.grid.major.x = element_blank(),
      panel.grid.minor.x = element_blank(),
      legend.position = "none",
      axis.text.x = element_text(angle = 45, hjust = 1, size = 12)
    ) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.1)))

  plot_list[[selected[i]]] <- p
}

n_plots <- length(plot_list)
n_cols  <- 5
n_rows  <- ceiling(n_plots / n_cols)

combined <- wrap_plots(plot_list, ncol = n_cols)
 
ggsave("Output/barplots/clot_formation_outlierless_v1.png",
       plot = combined,
       width = n_cols * 4,
       height = n_rows * 5,
       limitsize = FALSE)
```

### Fold change enrichment

```{r}
plot_enriched_barplot <- function(selected, title, outfile = NULL,
                                    x_font_size = 14, legend_pos = "right") {

  group_map <- list(
    "Plasma_EV" = grep("^P\\d{2}.*_log2FC$", colnames(protein_data_FC), value = TRUE),
    "Serum_EV"  = grep("^S\\d{2}.*_log2FC$", colnames(protein_data_FC), value = TRUE)
  )
  sample_cols <- unlist(group_map, use.names = FALSE)

  if (!dir.exists("Output/barplots")) dir.create("Output/barplots", recursive = TRUE)

  df_all <- protein_data_FC %>%
    filter(Protein.Accession %in% selected) %>%
    select(GeneName, Protein.Accession, all_of(sample_cols)) %>%
    pivot_longer(cols = all_of(sample_cols),
                 names_to = "sample",
                 values_to = "log2FC over total plasma or serum") %>%
    mutate(condition = case_when(
      sample %in% group_map[["Plasma_EV"]] ~ "Plasma EV",
      sample %in% group_map[["Serum_EV"]]  ~ "Serum EV"
    )) %>%
    filter(!is.na(`log2FC over total plasma or serum`))

  # preserve the order proteins appear in `selected`
  df_all$GeneName <- factor(df_all$GeneName,
                             levels = unique(df_all$GeneName[order(match(df_all$Protein.Accession, selected))]))
  df_all$condition <- factor(df_all$condition, levels = c("Plasma EV", "Serum EV"))

  p <- ggplot(df_all, aes(x = GeneName, y = `log2FC over total plasma or serum`,
                            fill = condition, group = condition)) +
    stat_summary(fun = mean,
                 geom = "bar",
                 color = "black",
                 position = position_dodge(width = 0.8),
                 width = 0.7) +
    stat_summary(fun.data = mean_sdl,
                 fun.args = list(mult = 1),
                 geom = "errorbar",
                 position = position_dodge(width = 0.8),
                 width = 0.3) +
    scale_fill_manual(values = c("Plasma EV" = "#9c5ae2", "Serum EV" = "#F2C438")) +
    theme_cowplot() +
    labs(
      title = title,
      y = "log2FC over total plasma or serum",
      x = NULL,
      fill = NULL
    ) +
    theme(
      panel.background = element_rect(fill = "white", color = NA),
      plot.background = element_rect(fill = "white", color = NA),
      panel.grid.major.x = element_blank(),
      panel.grid.minor.x = element_blank(),
      axis.text.x = element_text(angle = 45, hjust = 1, size = x_font_size),
      legend.position = legend_pos
    ) +
    scale_y_continuous(expand = expansion(mult = c(0.05, 0)))

  if (!is.null(outfile)) {
    ggsave(outfile, plot = p, width = max(8, length(selected) * 0.7), height = 5)
  }

  return(p)
}

p1 <- plot_enriched_barplot(Platelet$Uniprot.accession, "Platelet markers",
                              "Output/barplots/Platelet_FC_MS2_V5.pdf", legend_pos = c(0.1, 0.15))

p2 <- plot_enriched_barplot(Enriched$Uniprot.accession, "EV markers enriched by Mag-Net",
                              "Output/barplots/Enriched_FC_MS2_V5.pdf", legend_pos = c(0.2, 0.85))

p3 <- plot_enriched_barplot(Depleted$Uniprot.accession, "Markers depleted by Mag-Net",
                              "Output/barplots/Depleted_FC_MS2_V5.pdf", legend_pos = c(0.8, 0.15))
```

### Scatter plots for correlation

```{r}
plot_list <- list()

highlight_accessions <- Enriched$Uniprot.accession   # red
depleted_accessions  <- Depleted$Uniprot.accession    # blue -- swap in your actual depleted list

for (i in 1:20) {
  n <- sprintf("%02d", i)

  p_col <- grep(paste0("^P", n, "_log2FC"), colnames(protein_data_FC), value = TRUE)
  s_col <- grep(paste0("^S", n, "_log2FC"), colnames(protein_data_FC), value = TRUE)

  if (length(p_col) != 1 || length(s_col) != 1) {
    message("Skipping pair ", n, " — expected 1 match each, found P:", length(p_col), " S:", length(s_col))
    next
  }

  df <- protein_data_FC %>%
    select(GeneName, Protein.Accession, x = all_of(p_col), y = all_of(s_col)) %>%
    filter(!is.na(x), !is.na(y), !is.infinite(x), !is.infinite(y))

  if (nrow(df) == 0) {
    message("Skipping pair ", n, " — no complete points")
    next
  }

  r_all <- cor(df$x, df$y, use = "complete.obs")

  df <- df %>%
    mutate(
      quadrant = case_when(
        x >= 0 & y >= 0 ~ "Q1",
        x <  0 & y >= 0 ~ "Q2",
        x <  0 & y <  0 ~ "Q3",
        x >= 0 & y <  0 ~ "Q4"
      ),
      dist = sqrt(x^2 + y^2),
      is_highlighted = Protein.Accession %in% highlight_accessions,
      is_depleted    = Protein.Accession %in% depleted_accessions
    )

  top_labels <- df %>%
    group_by(quadrant) %>%
    slice_max(dist, n = 5, with_ties = FALSE) %>%
    ungroup()

  highlight_labels <- df %>% filter(is_highlighted)
  depleted_labels  <- df %>% filter(is_depleted)
  
combined_labels <- bind_rows(
  highlight_labels %>% mutate(label_color = "#D72638"),
  depleted_labels  %>% mutate(label_color = "#26567E")
)

p <- ggplot(df, aes(x = x, y = y)) +
    geom_point(color = "grey60", size = 1.5, alpha = 0.4) +
    geom_point(data = highlight_labels, color = "#D72638", size = 2, alpha = 1) +
    geom_point(data = depleted_labels, color = "#377EB8", size = 2, alpha = 1) +
    
    scale_color_identity() +
    geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
    geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
    geom_abline(slope = 1, linetype = "dashed") +
  geom_text_repel(data = combined_labels, aes(label = GeneName, color = label_color),
                     size = 2.5, fontface = "bold",
                     box.padding = 0.3, max.overlaps = 15,
                     show.legend = FALSE) +
    annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 1.5,
             label = paste0("R = ", round(r_all, 2)),
             color = "grey40", size = 3, fontface = "bold") +
    labs(title = paste0("Pair ", n), x = NULL, y = NULL) +
    coord_cartesian(xlim = c(-10, 10), ylim = c(-10, 10)) +
    theme_cowplot() +
    theme(plot.title = element_text(size = 11),
          panel.border = element_rect(color = "black", fill = NA, linewidth = 1))

  plot_list[[n]] <- p
}

combined <- wrap_plots(plot_list, ncol = 5)

ggsave("Output/scatterplots/paired_PS_grid_enriched_dep_V2.pdf",
       plot = combined,
       width = 5 * 3,
       height = 4 * 3,
       limitsize = FALSE)
```

```{r}
library(ComplexHeatmap)
library(circlize)
library(grid)
library(dplyr)

if (!dir.exists("Output")) dir.create("Output", recursive = TRUE)

## --- 1. Select the log2FC columns for each condition -----------------

plasma_log2fc_cols <- grep("^P\\d{2}_.*_log2FC$", colnames(protein_data_FC), value = TRUE)
serum_log2fc_cols  <- grep("^S\\d{2}_.*_log2FC$", colnames(protein_data_FC), value = TRUE)

df_plasma <- protein_data_FC %>%
  select(all_of(plasma_log2fc_cols)) %>%
  filter(if_all(everything(), ~ !is.infinite(.))) %>%
  na.omit()

df_serum <- protein_data_FC %>%
  select(all_of(serum_log2fc_cols)) %>%
  filter(if_all(everything(), ~ !is.infinite(.))) %>%
  na.omit()

df_both <- protein_data_FC %>%
  select(all_of(c(plasma_log2fc_cols, serum_log2fc_cols))) %>%
  filter(if_all(everything(), ~ !is.infinite(.))) %>%
  na.omit()

## --- 2. Compute pairwise Spearman correlations ------------------------

corrs_plasma <- cor(df_plasma, method = "spearman")
corrs_serum  <- cor(df_serum,  method = "spearman")
corrs_all    <- cor(df_both,   method = "spearman")

## --- 3. Heatmaps -------------------------------------------------------

col_fun <- colorRamp2(c(0.6, 0.8, 1), c("blue", "white", "red"))ß

plot_corr_heatmap <- function(mat, filename, width = 12, height = 12) {
  pdf(filename, width = width, height = height)
  ht <- Heatmap(
    mat,
    cluster_rows = TRUE,
    cluster_columns = TRUE,
    show_row_names = TRUE,
    show_column_names = TRUE,
    column_names_side = "top",
    col = col_fun,
    cell_fun = function(j, i, x, y, width, height, fill) {
      grid.text(
        round(mat[i, j], 3),
        x, y,
        gp = gpar(fontsize = 8)
      )
    }
  )
  draw(ht)
  dev.off()
}

plot_corr_heatmap(corrs_plasma, "Output/log2FC_plasma_corr.pdf")
plot_corr_heatmap(corrs_serum,  "Output/log2FC_serum_corr.pdf")
plot_corr_heatmap(corrs_all,    "Output/log2FC_combined_corr.pdf", width = 18, height = 18)
```

### Scatter plot, on average

```{r}
highlight_accessions <- Enriched$Uniprot.accession   # red
depleted_accessions  <- Depleted$Uniprot.accession    # blue
Coag_accessions <- Coagulation$Uniprot.accession

df <- protein_data_FC %>%
  mutate(
    x = log2(Plasma_EV_avg / Plasma_total_avg),
    y = log2(Serum_EV_avg / Serum_total_avg) 
  ) %>%
  select(GeneName, Protein.Accession, x, y) %>%
  filter(!is.na(x), !is.na(y), !is.infinite(x), !is.infinite(y))

r_all <- cor(df$x, df$y, use = "complete.obs")

df <- df %>%
  mutate(
    quadrant = case_when(
      x >= 0 & y >= 0 ~ "Q1",
      x <  0 & y >= 0 ~ "Q2",
      x <  0 & y <  0 ~ "Q3",
      x >= 0 & y <  0 ~ "Q4"
    ),
    dist = sqrt(x^2 + y^2),
    is_highlighted = Protein.Accession %in% highlight_accessions,
    is_depleted    = Protein.Accession %in% depleted_accessions,
    is_coag = Protein.Accession %in% Coag_accessions
  )

top_labels <- df %>%
  group_by(quadrant) %>%
  slice_max(dist, n = 5, with_ties = FALSE) %>%
  ungroup()

highlight_labels <- df %>% filter(is_highlighted)
depleted_labels  <- df %>% filter(is_depleted)
coag_labels <- df %>% filter(is_coag)

combined_labels <- bind_rows(
  highlight_labels %>% mutate(label_color = "#D72638"),
  depleted_labels  %>% mutate(label_color = "#377EB8"),
  coag_labels %>% mutate(label_color = "#EC9F05")
  
)

p <- ggplot(df, aes(x = x, y = y)) +
  geom_point(color = "grey60", size = 1.5, alpha = 0.4) +
  geom_point(data = highlight_labels, color = "#D72638", size = 2, alpha = 1) +
  geom_point(data = depleted_labels, color = "#377EB8", size = 2, alpha = 1) +
  geom_point(data = coag_labels, color = "#EC9F05", size = 2, alpha = 1) +
  scale_color_identity() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  geom_abline(slope = 1, linetype = "dashed") +
  geom_text_repel(data = combined_labels, aes(label = GeneName, color = label_color),
                 size = 4, fontface = "bold",
                 box.padding = 0.6,        # more space around each label
                 point.padding = 0.5,      # more space between label and its point
                 force = 3,                # stronger repulsion between labels
                 force_pull = 0.5,         # weaker pull back toward the point (lets them drift further)
                 max.overlaps = 20,
                 show.legend = FALSE,
                 bg.color = "white", bg.r = 0.2,
                 seed = 42) +              # keep layout reproducible while you tune
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 1.5,
           label = paste0("R = ", round(r_all, 2)),
           color = "grey40", size = 3, fontface = "bold") +
  labs(title = "Average log2(EV/total)", x = "Plasma log2(EV/plasma total)", y = "Serum log2(EV/serum total)") +
  coord_cartesian(xlim = c(-10, 10), ylim = c(-10, 10)) +
  theme_cowplot() +
  theme(plot.title = element_text(size = 11),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 1))

ggsave("Output/scatterplots/average_PS_enriched_dep.pdf",
       plot = p,
       width = 6,
       height = 6)
```

```{r}
T1D_samples <- read.csv("Output/251125_normed_proteins.csv")

selected <- Enriched$Uniprot.accession

group_map <- list(
  "Tocil"  = grep("^T\\d{2}_", colnames(T1D_samples), value = TRUE),
  "Siltux" = grep("^S\\d{2}_", colnames(T1D_samples), value = TRUE)
)

sample_cols <- unlist(group_map, use.names = FALSE)

# Helper: extract the 2-digit sample number and determine odd/even -> pre/post
get_timepoint <- function(sample_name) {
  num <- as.integer(sub("^[TS](\\d{2})_.*", "\\1", sample_name))
  ifelse(num %% 2 == 1, "pre", "post")
}

if (!dir.exists("Output/barplots")) dir.create("Output/barplots", recursive = TRUE)

library(patchwork)

plot_list <- list()

for (i in seq_along(selected)) {

  df_plot <- T1D_samples %>%
    filter(Protein.Accession == selected[i])

  if (nrow(df_plot) == 0) {
    message("Skipping missing protein: ", selected[i])
    next
  }

  df_long <- df_plot %>%
    select(GeneName, all_of(sample_cols)) %>%
    pivot_longer(cols = all_of(sample_cols),
                 names_to = "sample",
                 values_to = "normalized abundance") %>%
    mutate(
      drug = case_when(
        sample %in% group_map[["Tocil"]]  ~ "Tocil",
        sample %in% group_map[["Siltux"]] ~ "Siltux"
      ),
      timepoint = get_timepoint(sample),
      condition = paste(drug, timepoint, sep = "_")
    ) %>%
    filter(!is.na(`normalized abundance`))

  if (nrow(df_long) == 0) {
    message("Skipping protein with no non-missing values: ", selected[i])
    next
  }

  df_long$condition <- factor(df_long$condition,
                               levels = c("Tocil_pre", "Tocil_post", "Siltux_pre", "Siltux_post"))

  p <- ggplot(df_long, aes(x = condition, y = `normalized abundance`, fill = condition)) +
    stat_summary(fun = mean,
                 geom = "bar",
                 color = "black",
                 width = 0.7,
                 alpha = 0.8) +
    stat_summary(fun.data = mean_sdl,
                 fun.args = list(mult = 1),
                 geom = "errorbar", width = 0.2) +
    geom_quasirandom(width = 0.2, size = 2, shape = 21, color = "black") +
    scale_fill_manual(values = c("Tocil_pre" = "#89a9e4", 
                                 "Tocil_post" = "#0a369d", 
                                 "Siltux_pre" = "#f192aa", 
                                 "Siltux_post"= "#da4167")) +
    theme_cowplot() +
    labs(
      title = df_long$GeneName[[1]],
      y = "normalized abundance",
      x = NULL
    ) +
    theme(
      panel.background = element_rect(fill = "white", color = NA),
      plot.background = element_rect(fill = "white", color = NA),
      panel.grid.major.x = element_blank(),
      panel.grid.minor.x = element_blank(),
      legend.position = "none",
      axis.text.x = element_text(angle = 45, hjust = 1, size = 10)
    ) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.1)))

  plot_list[[selected[i]]] <- p
}

n_plots <- length(plot_list)
n_cols  <- 5
n_rows  <- ceiling(n_plots / n_cols)

combined <- wrap_plots(plot_list, ncol = n_cols)

ggsave("Output/barplots/Enriched_T1D_MS2_4bar.png",
       plot = combined,
       width = n_cols * 4,
       height = n_rows * 5,
       limitsize = FALSE)
```

### PCA

```{r}


sample_cols <- c(
  grep("^P\\d{2}", colnames(protein_data), value = TRUE),
  grep("^S\\d{2}", colnames(protein_data), value = TRUE),
  grep("^PlasmaTotal", colnames(protein_data), value = TRUE),
  grep("^SerumTotal", colnames(protein_data), value = TRUE)
)

protein_data_matrix <- protein_data %>%
  select(GeneName, all_of(sample_cols)) %>%
  filter(if_all(all_of(sample_cols), ~ !is.na(.) & !is.infinite(.))) 

gene_names <- protein_data_matrix$GeneName

mat <- protein_data_matrix %>%
  select(all_of(sample_cols)) %>%
  as.matrix()

rownames(mat) <- gene_names

## --- 2. Run PCA on samples (rows = samples after transpose) -----------

PCA <- prcomp(t(mat), scale. = TRUE)
PCA_x <- as.data.frame(PCA$x)
PCA_x$sample <- rownames(PCA_x)

## --- 3. Assign group labels for coloring -------------------------------

PCA_x <- PCA_x %>%
  mutate(group = case_when(
    grepl("^P\\d{2}", sample) ~ "Plasma_EV",
    grepl("^S\\d{2}", sample) ~ "Serum_EV",
    grepl("^PlasmaTotal", sample) ~ "Plasma_total",
    grepl("^SerumTotal", sample) ~ "Serum_total"
  ))

## --- 4. Scores plot -----------------------------------------------------

samples_to_label <- c("P03", "P04")  # swap in your actual sample names

label_df <- PCA_x %>% filter(sample %in% samples_to_label)

var_explained <- round(summary(PCA)$importance[2,] * 100, 1 )

ggplot(PCA_x, aes(x = PC1, y = PC2, color = group)) +
  geom_point(size = 4, stroke = 0) +
  geom_text_repel(data = label_df, aes(label = sample),
                   color = "black", size = 3, fontface = "bold",
                   box.padding = 0.5, max.overlaps = Inf) +
  scale_color_manual(values = c(
    "Plasma_EV"    = "#9c5ae2",
    "Serum_EV"     = "#F2C438",
    "Plasma_total" = "#377EB8",
    "Serum_total"  = "#D72638"
  )) +
  theme_cowplot() +
  theme(
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
    legend.position = c(0.98, 0.98),
    legend.justification = c("right", "top"),
    legend.background = element_rect(fill = "white", color = "black", linewidth = 0.4),
    legend.margin = margin(4, 6, 4, 6)
  ) +
  labs(
       x = paste0("PCA1 (", var_explained[1], "%)"),
       y = paste0("PCA2 (", var_explained[2], "%)"),
       color = NULL)
ggsave("output/SVP_PCA_v2.PDF", width=4, height=3.5)

```
